---
title: "Fig4_np_mussel_feeding_exp"
author: "Yu-Cheng Marvin Lin"
date: "06-05-26"
output: html_document
---

```{r setup}
library(rstatix)
library(ggpubr)
library(tidyverse)
library(lubridate)
library(glue)
library(readxl)
library(googlesheets4)
library(broom.mixed)
library(openxlsx)
source("functions.R")
```

# Objectives

This script analyzes the result of a quagga mussel feeding experiment. The goal of the experiment was to test the hypothesis that the feeding outcome between *Microcystis aeruginosa* and quagga mussels is dependent on the elemental ratios of nutrient supply, and that this dependence is strain-specific. We chose two unpalatable and one inhibitory *Microcystis* strains (LE19-131, LE19-195 and LE19-114) and mixed them with the preferred food - green alga *Chlamydomonas* sp. as food mixtures subjected to mussel grazing for 1.5 hrs under three different N:P ratios of nutrient supply (5:1, 20:1, 80:1). These ratios represent the lowered and elevated ratios comparing to the 20:1 ratio at which we conducted our strain-based feeding experiment. We measured the clearance and ingestion rates of quagga mussels on both phytoplankton species using a fluoroprobe.

# Analysis

## Data preparation

### Import fluoroprobe data

```{r warning=FALSE, message = FALSE}
#Make a list of the paths of files so that R can import with read_tsv
file_list <- list.files(path = "inputs/formatted_rawdata_fp",
                        full.names = TRUE, ignore.case = TRUE, pattern = "QMF_.*txt$")
raw_data_list <- list()
for (file in seq_along(file_list)){
  fp_readings <- read_tsv(file_list[file], col_names = TRUE) %>%
    filter(Culture != 'MYCO') %>% # remove one pilot data
    mutate(exp_date= basename(file_list[file]) %>%
             str_replace("QMF_","X")%>%
             str_remove_all("\\.txt")) %>%
    # make NPratio factor
    mutate(NPratio = as.factor(NPratio)) %>%
    drop_na()
  raw_data_list[[file]] <- fp_readings
}
raw_data <- bind_rows(raw_data_list) %>%
  # remove rows containing NAs in any columns invalid multibyte string
  filter(if_all(everything(), ~ !is.na(.))) %>%
  # assign data type to each column
  mutate(across(c(`Bluegreen`, `Green Algae`, `Diatoms`, `Cryptophyta`, `Yellow substances`, `Total conc.`), as.numeric)) %>%
  mutate(across(-c(`Bluegreen`, `Green Algae`, `Diatoms`, `Cryptophyta`, `Yellow substances`, `Total conc.`), as.factor))

str(raw_data)
```

### Filter out failed experiments and invalid readings with reasons
```{r}
# Remove failed experiments and invalid readings
raw_data <- raw_data %>%
  # remove exp on 10/24/25 because there were bluegreen contamination in the mussel tank that got carried over to the experimental beakers.
  filter(!(exp_date == 'X102425')) %>%
  # remove the mussels that closed and stopped feeding after a period of time. we removed them from the beakers after they are not feeding anymore
  filter(!(exp_date == 'X111425' & Treatment == "MIX3_Q3_80" & Timepoint == "T5")) %>%
  # remove zero or negative readings. Fluoroprobe pick up some readings as zero or negative out of the 7 readings at a time.
  filter(!( `Total conc.` <= 0))
```


### Summairize fluoroprobe data

The fluoroprobe took multiple (n = 7-10) readings at a time from the same sample. We want to take the average per beaker per timepoint.

```{r}
#Import fluoroprobe data from mussel feeding exp
fp_data <-  raw_data %>%
  rename("blugrn" = Bluegreen,
         "time" = `Date/Time`,
         "grn" = `Green Algae`,
         "total_chla" = `Total conc.`) %>%
  # Sometimes chlamy readings split to other fractions when no bluegreen is present in the control beakers
  # replace grn readings with total conc in cases where it is chlamy control beakers
  #mutate(grn = case_when(Culture == 'CHLAMY' & GROUP == 'CONTROL' ~ total_chla,
  #                       T ~ grn)) %>%
  # make sure time column is in date-time format
  mutate(time = mdy_hm(time)) %>%
  mutate(dd = lubridate::decimal_date(time),
         Date = format(time,"%m-%d-%Y")) %>%
  # calculate how much percentage is split into other fractions
  mutate(perc_error = (blugrn+grn)/total_chla) %>%
  pivot_longer(cols = c(blugrn,grn,total_chla), 
               names_to = 'category', values_to = 'chla') %>%
  # make sure chla is factor
  mutate(category = factor(category, levels = c('grn','blugrn','total_chla'))) %>%
  mutate(chla = as.numeric(chla)) 
  
  

## The beaker list, including all the combinations of culture, nutrient ratio, replicate, and group tested
metadata <- read_tsv('inputs/QMF_metadata_2025.txt', col_names = T) %>%
  mutate(across(c(`GROUP`,`REP`,`NPratio`,`NOTE`,`mussel_batch`),as.factor)) %>%
  mutate(Culture = as.character(`Culture`)) %>%
  # make sure time column is in date-time format
  mutate(start_time = mdy_hm(start_time)) 

## Summarize the fluoroprobe readings on a per beaker basis
fp_summary <- fp_data %>%
  group_by(Date, Culture, GROUP, REP, NPratio, NOTE, Timepoint, category, exp_date) %>%
  summarise(mean_chla = mean(chla, na.rm = TRUE),
            mean_time = mean(time, na.rm = TRUE),
            mean_perc_error = mean(perc_error, na.rm = TRUE),
            sd_chla = sd(chla, na.rm = TRUE),
            sd_time = sd(time, na.rm = TRUE),
            sd_perc_error = sd(perc_error, na.rm = TRUE),
            n = n()) %>%
  ungroup()

## Check if all the metadata are included
wm <- fp_summary %>% 
  filter(Timepoint == 'T0') %>%
  group_by(Culture, GROUP, REP, NPratio, NOTE, exp_date) %>%
  summarise(n = n()) %>%
  full_join(metadata %>% select(-c('start_time','m1_body_length','m2_body_length','volume_ml','mussel_batch')), by = c('Culture', 'GROUP', 'REP', 'NPratio', 'NOTE', 'exp_date'), keep = FALSE) %>%
  # show the non-matching rows
  filter(if_any(everything(), ~ is.na(.)))
```

### Time Correction and combining the data

It seems like we're missing one T0 reading of MIX2_Q1_20 from 08/15/25.

Because mussels were added in the beakers between T0 and T1, we need to correct the T0 time to reflect the actual start time of the experiment. Therefore, we correct the T0 to the time we observed mussel extended their siphons.

```{r T0 correction}
# Correct T0 by replacing with documented quagga mussel openning time
t0_correct <- fp_summary %>%
  filter(Timepoint %in% c("T0","T1")) %>%
  group_by(Date, Culture, GROUP, REP, NPratio, NOTE, category) %>%
  filter(n() > 1) %>% #There are a couple of timepoints without a T0
  left_join(metadata %>% select(c('Culture', 'GROUP', 'REP', 'NPratio', 'NOTE', 'exp_date', 'start_time')), 
            by = c('Culture', 'GROUP', 'REP', 'NPratio', 'NOTE', 'exp_date'), keep = FALSE) %>%
  filter(Timepoint == 'T0') %>%
  mutate(mean_time = case_when(Culture == 'QUAGGA' ~ start_time,
                               T ~ mean_time)) %>%
  ungroup() %>%
  select(- c(start_time))
# Check these edge cases
addt0 <- fp_summary %>%
  filter(Timepoint %in% c("T0","T1")) %>%
  group_by(Date, Culture, GROUP, REP, NPratio, NOTE, category) %>%
  filter(n() <= 1) %>%
  ungroup()

# Since one is control, we can use the average T0 time from the same date and culture
# another one is missing T0 for mix2, we can add it back without correction
t0_correct <- rbind(t0_correct, addt0 %>% filter(Timepoint == 'T0'))

# Combine the corrected T0 times with the rest of the data, and make the data ready for analysis
full_data <- rbind(filter(fp_summary,Timepoint != 'T0'), t0_correct) %>%
  group_by(Date, Culture, GROUP, REP, NPratio, NOTE, category) %>%
  # calculate time intervals between readings
  mutate(
    begin = min(mean_time),
    read_interval = interval(ymd_hms(begin), ymd_hms(mean_time)),
    hours = int_length(read_interval) / as.numeric(dhours(1)),
    #Get time between each read
    source = case_when(Timepoint == "TF" ~ "Mixed", #Water column is during the experiment while mixed is the final reading.
                       T ~ "Water Column")) %>%
  # incorporate mussel biomass data and beaker volume data
  left_join(metadata %>% select(-c('start_time')), by = c('Culture', 'GROUP', 'REP', 'NPratio', 'NOTE', 'exp_date'), keep = FALSE) %>%
  # calculate the volume of water in the beaker during the experiment. This is important for clearance rate calculation
  mutate(# volume of water removed
         VFWC = case_when(Timepoint == 'T0' ~ 25,
                          Timepoint == 'T1' ~ 50,
                          Timepoint == 'T2' ~ 75,
                          Timepoint == 'T3' ~ 100,
                          Timepoint == 'T4' ~ 125,
                          Timepoint == 'T5' ~ 150,
                          Timepoint == 'TF' ~ 175),
         # Final volume of water in the beaker
         VFM = volume_ml - VFWC)
```

## Clearance Rate calculation

### Accounting for mussel biomass

The size and weight of mussel is an important factor for their clearance rate. We built a linear model for estimating mussel ash-free dry weight (AFDW) from shell length, based on the mussel dissection data. We will use this data to estimate the AFDW and get adjusted clearance rate per beaker.

```{r Mussel AFDW}
# Import mussel dissection data
mussel_info <- read_csv("inputs/mussel_DW_npexp.csv") %>%
  mutate(GDW = gut_dry_mass - capsule_mass,
         PAW = gut_pa_mass - capsule_mass,
         AFDW = GDW - PAW,
         combust_eff = AFDW/GDW,
         batch = as.factor(batch)) %>%
  # a high combustion efficiency indicates 
  filter(combust_eff > 0.79)

#Linear regression of AFDW and GDW of mussel tissue and shell-length
AFDW_lm <- lm(AFDW ~ mussel_length, data = mussel_info)
GDW_lm <- lm(GDW ~ mussel_length, data = mussel_info)
summary(AFDW_lm) #Adjusted R^2 0.7969 3.85e-07
summary(GDW_lm) #Adjusted R^2 0.8064 2.602e-07

mussel_info$AFDW_lm <- AFDW_lm$fitted.values
mussel_info$GDW_lm <- GDW_lm$fitted.values

#Check for normality
shapiro_test(mussel_info$AFDW)
shapiro_test(mussel_info$GDW)
shapiro_test(mussel_info$mussel_length)
ggqqplot(mussel_info$mussel_length)

#Check residuals vs fitted plot
plot(lm(AFDW ~ mussel_length, data = mussel_info))
plot(lm(GDW ~ mussel_length, data = mussel_info))
plot(lm(GDW ~ PAW, data = mussel_info))
```

These data are from two different batches of mussels, WE4 and WE6.

```{r}
AFDW_plot <- ggplot(mussel_info, aes(x = mussel_length, y = AFDW)) +
  geom_point(color = "black", size = 3) +
  geom_smooth(method = "lm",
              formula = y~x, level = 0.95, alpha = .15) +
  labs(y = "Ash-Free Dry Weight (mg)",
       x = "Mussel Shell Length (mm)") +
  stat_regline_equation(label.y = c(19,18)) +
  stat_cor(method = "pearson",
    aes(label = paste(after_stat(rr.label), after_stat(p.label), sep = "~`,`~"))) + 
  theme_pubr(base_size = 14)

GDW_plot <- ggplot(mussel_info, aes(x = mussel_length, y = GDW)) +
  geom_point(color = "black", size = 3) +
  geom_smooth(method = "lm",
              formula = y~x, level = 0.95, alpha = .15) +
  labs(y = "Mussel Tissue Dry Weight (mg)",
       x = "Mussel Shell Length (mm)") +
  stat_regline_equation(label.y = 21) +
  stat_cor(method = "pearson",
    aes(label = paste(after_stat(rr.label), after_stat(p.label), sep = "~`,`~"))) + 
  theme_pubr(base_size = 14)
AFDW_plot;GDW_plot
```

Standardize clearance rate to mussel AFDW

```{r}
# Fit model to estimate AFDW from mussel shell length
feeding_data <- full_data %>%
  group_by(Date, Culture, GROUP, REP, NPratio, NOTE, Timepoint, category) %>% 
  mutate(M1AFDW = -7.5 + 1.1*m1_body_length,
         M2AFDW = -7.5 + 1.1*m1_body_length,
         total_AFDW = case_when(is.na(M1AFDW) ~ M2AFDW,
                                is.na(M2AFDW) ~ M1AFDW,
                                T ~ (M1AFDW + M2AFDW)),
         # Convert AFDW to carbon content assuming 52.4% of AFDW is carbon (from mussel dissection data)
         total_C = total_AFDW * 0.524, 
         percent_C = total_C/total_AFDW) %>%
  ungroup()
```

### Calculate Clearance and Ingestion Rates

#### Control data: to account for settling and death

```{r}
#Calculations for control data to get final water column and mixed measurements
control_data <- feeding_data %>%
  filter(GROUP == 'CONTROL') %>%
  select(Date, Culture, GROUP, REP, NPratio, NOTE, Timepoint, category, 
         mean_chla, mean_time, hours, volume_ml, VFM, VFWC)

# examine chlamy control from a boxplot
chlamy_control <- ggplot(control_data %>% filter(Culture == 'CHLAMY'), 
                         aes(x = Timepoint, y = mean_chla)) +
  # plot two boxplot per timepoint one for each category
  geom_boxplot(aes(fill = category), position = position_dodge(width = 0.75), outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Date), position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75), shape = 21, size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("blugrn" = "cadetblue",
                               "grn" = "yellowgreen", 
                               "total_chla" = "plum4"),
                                labels = c("blugrn" = "*M. aeruginosa*",
                                           "grn" = "*C. oblonga*",
                                           "total_chla" = "Total phytoplanktons"), 'category') +
  facet_wrap(~NPratio, nrow = 1, labeller = labeller(NPratio = c("5" = "N:P 5:1", "20" = "N:P 20:1", "LOW20" = "N:P 20:1 at lower conc.", "40" = "N:P 40:1", "80" = "N:P 80:1"))) +
  scale_y_continuous(limits = c(0, 4)) +
  labs(y = expression("Chl a ("*mu*"g/L)"),
       x = "Timepoint") +
  theme_classic()+
  theme(legend.position = 'right')
#ggsave(chlamy_control, filename = "control_chlamy_boxplot.png", width = 8, height = 3, units = "in", dpi = 300)
# check mix1 control (microcystis strain == 19-195)
mix1_control <- ggplot(control_data %>% filter(Culture == 'LE19-195'), 
                         aes(x = Timepoint, y = mean_chla)) +
  # plot two boxplot per timepoint one for each category
  geom_boxplot(aes(fill = category), position = position_dodge(width = 0.75), outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Date), position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75), shape = 21, size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("blugrn" = "cadetblue",
                               "grn" = "yellowgreen", 
                               "total_chla" = "plum4"),
                                labels = c("blugrn" = "*M. aeruginosa*",
                                           "grn" = "*C. oblonga*",
                                           "total_chla" = "Total phytoplanktons"), 'category')+
  facet_wrap(~NPratio, nrow = 1, labeller = labeller(NPratio = c("5" = "N:P 5:1", "20" = "N:P 20:1", "LOW20" = "N:P 20:1 at lower conc.", "40" = "N:P 40:1", "80" = "N:P 80:1"))) +
  scale_y_continuous(limits = c(0, 4)) +
  labs(
    title = "LE19-195 (Mix 1) Control",
    y = expression("Chl a ("*mu*"g/L)"),
    x = "Timepoint") +
  theme_classic()+
  theme(legend.position = 'right')
#ggsave(mix1_control, filename = "control_LE19195_boxplot.png", width = 8, height = 3, units = "in", dpi = 300)
# check mix2 control (microcystis strain == 19-131)
mix2_control <- ggplot(control_data %>% filter(Culture == 'LE19-131'), 
                         aes(x = Timepoint, y = mean_chla)) +
  # plot two boxplot per timepoint one for each category
  geom_boxplot(aes(fill = category), position = position_dodge(width = 0.75), outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Date), position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75), shape = 21, size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("blugrn" = "cadetblue",
                               "grn" = "yellowgreen", 
                               "total_chla" = "plum4"),
                                labels = c("blugrn" = "*M. aeruginosa*",
                                           "grn" = "*C. oblonga*",
                                           "total_chla" = "Total phytoplanktons"), 'category')+
  facet_wrap(~NPratio, nrow = 1, labeller = labeller(NPratio = c("5" = "N:P 5:1", "20" = "N:P 20:1", "LOW20" = "N:P 20:1 at lower conc.", "40" = "N:P 40:1", "80" = "N:P 80:1"))) +
  scale_y_continuous(limits = c(0, 4)) +
  labs(
    title = "LE19-131 (Mix 2) Control",
    y = expression("Chl a ("*mu*"g/L)"),
    x = "Timepoint") +
  theme_classic()+
  theme(legend.position = 'right')
#ggsave(mix2_control, filename = "control_LE19131_boxplot.png", width = 8, height = 3, units = "in", dpi = 300)
# check mix3 control (microcystis strain == 19-114)
mix3_control <- ggplot(control_data %>% filter(Culture == 'LE19-114'), 
                         aes(x = Timepoint, y = mean_chla)) +
  # plot two boxplot per timepoint one for each category
  geom_boxplot(aes(fill = category), position = position_dodge(width = 0.75), outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Date), position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75), shape = 21, size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("blugrn" = "cadetblue",
                               "grn" = "yellowgreen", 
                               "total_chla" = "plum4"),
                                labels = c("blugrn" = "*M. aeruginosa*",
                                           "grn" = "*C. oblonga*",
                                           "total_chla" = "Total phytoplanktons"), 'category')+
  facet_wrap(~NPratio, nrow = 1, labeller = labeller(NPratio = c("5" = "N:P 5:1", "20" = "N:P 20:1", "LOW20" = "N:P 20:1 at lower conc.", "40" = "N:P 40:1", "80" = "N:P 80:1"))) +
  scale_y_continuous(limits = c(0, 4)) +
  labs(
    title = "LE19-114 (Mix 3) Control",
    y = expression("Chl a ("*mu*"g/L)"),
    x = "Timepoint") +
  theme_classic()+
  theme(legend.position = 'right')
#ggsave(mix3_control, filename = "control_LE19114_boxplot.png", width = 8, height = 3, units = "in", dpi = 300)
chlamy_control;mix1_control;mix2_control;mix3_control
```

Calculate the final chla concentration in the control beakers, corrected for the volume removed for sampling. This will be the baseline growth control for clearance and ingestion rates calculation in the quagga beakers.

```{r}
# Calculation for growth and death rate
grn_summary <- control_data %>%
  filter(category == 'grn') %>% 
  select(Date, Timepoint, Culture, GROUP, NPratio, NOTE, REP, mean_chla) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_chla) %>%
  mutate(
    # let C0 be the initial concentration at T0 or the second lowest timepoint
    C0 = ifelse(is.na(T0), T1, T0),
    #Final control measurement before mixing
    CFWC = T5, 
    #Final control measurement after mixing
    CFM = TF) %>%
  select(-c(T0, T1, T2, T3, T4, T5, TF)) %>%
  left_join(control_data %>% filter(category == 'grn') %>%select(-c(mean_chla,mean_time,category)), 
            by = c('Date', 'Culture', 'GROUP', 'NPratio', 'NOTE', 'REP'), keep = FALSE) %>%
  #Control final corrected for sample volume removed
  filter(Timepoint == 'TF') %>% #we want the final volume
  mutate(CFMx = (VFM/volume_ml)*CFM+(VFWC/volume_ml)*CFWC) %>%
  # create a label column for easier handling
  mutate(label = paste(Culture, GROUP, NPratio, sep = "_")) %>%
  mutate(category = 'grn') %>%
  select(-c(Timepoint, VFM, VFWC, volume_ml))

blugrn_summary <- control_data %>%
  filter(category == 'blugrn') %>% 
  select(Date, Timepoint, Culture, GROUP, NPratio, NOTE, REP, mean_chla) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_chla) %>%
  mutate(
    # let C0 be the initial concentration at T0 or the second lowest timepoint
    C0 = ifelse(is.na(T0), T1, T0),
    #Final control measurement before mixing
    CFWC = T5, 
    #Final control measurement after mixing
    CFM = TF) %>%
  select(-c(T0, T1, T2, T3, T4, T5, TF)) %>%
  left_join(control_data %>% filter(category == 'blugrn') %>%select(-c(mean_chla,mean_time,category)), 
            by = c('Date', 'Culture', 'GROUP', 'NPratio', 'NOTE', 'REP'), keep = FALSE) %>%
  #Control final corrected for sample volume removed
  filter(Timepoint == 'TF') %>% #we want the final volume
  mutate(CFMx = (VFM/volume_ml)*CFM+(VFWC/volume_ml)*CFWC) %>%
  # create a label column for easier handling
  mutate(label = paste(Culture, GROUP, NPratio, sep = "_")) %>%
  mutate(category = 'blugrn') %>%
  select(-c(Timepoint, VFM, VFWC, volume_ml))

total_chla_summary <- control_data %>%
  filter(category == 'total_chla') %>% 
  select(Date, Timepoint, Culture, GROUP, NPratio, NOTE, REP, mean_chla) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_chla) %>%
  mutate(
    # let C0 be the initial concentration at T0 or the second lowest timepoint
    C0 = ifelse(is.na(T0), T1, T0),
    #Final control measurement before mixing
    CFWC = T5, 
    #Final control measurement after mixing
    CFM = TF) %>%
  select(-c(T0, T1, T2, T3, T4, T5, TF)) %>%
  left_join(control_data %>% filter(category == 'total_chla') %>%select(-c(mean_chla,mean_time,category)), 
            by = c('Date', 'Culture', 'GROUP', 'NPratio', 'NOTE', 'REP'), keep = FALSE) %>%
  #Control final corrected for sample volume removed
  filter(Timepoint == 'TF') %>% #we want the final volume
  mutate(CFMx = (VFM/volume_ml)*CFM+(VFWC/volume_ml)*CFWC) %>%
  # create a label column for easier handling
  mutate(label = paste(Culture, GROUP, NPratio, sep = "_")) %>%
  mutate(category = 'total_chla') %>%
  select(-c(Timepoint, VFM, VFWC, volume_ml))

control_summary <- bind_rows(grn_summary, blugrn_summary,total_chla_summary)

# visualize the summary
ctrl_plot <- ggplot(control_summary, aes(x = NPratio, y = CFMx)) +
  geom_boxplot(aes(fill = category), position = position_dodge(width = 0.5), outlier.shape = NA, alpha = 0.5) +
  geom_point(aes(color = Date), size = 1.5, position = position_dodge(width = 0.5)) +
  scale_fill_manual(values = c("blugrn" = "cadetblue",
                               "grn" = "yellowgreen",
                               "total_chla" = "plum4"),
                                labels = c("blugrn" = "*M. aeruginosa*",
                                           "grn" = "*C. oblonga*",
                                           "total_chla" = "Total phytoplanktons"), 'category') +
  facet_wrap(~Culture, nrow = 1, 
             labeller = labeller(Culture = c("CHLAMY" = "Chlamydomonas oblonga",
                                                            
                                             "LE19-195" = "Mix 1: *M. aeruginosa* LE19-195",
                                                            
                                             "LE19-131" = "Mix 2: *M. aeruginosa* LE19-131",
                                             "LE19-114" = "Mix 3: *M. aeruginosa* LE19-114"))) +
  labs(title = 'Volume-corrected TF Chl a Concentration in Control Beakers',
    y = expression("Chl a ("*mu*"g/L)"),
    x = "Control Beakers (Culture_NPratio)") +
  theme_bw() +
  theme(legend.position = 'right') +
  rotate_x_text(angle = 60)
ctrl_plot
```

Calculate the final chla concentration in the experimental beakers, corrected for the volume removed for sampling. This will be used to calculate the clearance and ingestion rates in the quagga beakers.

```{r}
# Calculations for experimental data to get final water column and mixed measurements
mix_data <- feeding_data %>%
  filter(GROUP == 'QUAGGA') %>%
  select(Date, Culture, GROUP, REP, NPratio, NOTE, Timepoint, category, 
         mean_chla, mean_time, hours, volume_ml, VFM, VFWC)
## Process green readings
grn_mix_summary <- mix_data %>%
  filter(category == 'grn') %>% 
  select(Date, Timepoint, REP, Culture, GROUP, NPratio, NOTE, mean_chla) %>%
  group_by(Date, Timepoint, REP, Culture, GROUP, NPratio, NOTE) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_chla) %>%
  mutate(
  # let Z0 be the initial concentration at T0 or the second lowest timepoint
    Z0 = ifelse(is.na(T0), T1, T0),
    #Final control measurement before mixing
    ZFWC = ifelse(is.na(T5), T4, T5),
    #Final control measurement after mixing
    ZFM = TF) %>%
  select(-c(T0, T1, T2, T3, T4, T5, TF)) %>%
  left_join(mix_data %>% 
              filter(category == 'grn') %>%
              select(-c(mean_chla,mean_time,category)), 
            by = c('Date', 'REP', 'Culture', 'GROUP', 'NPratio', 'NOTE'), keep = FALSE) %>%
  #Control final corrected for sample volume removed
  filter(Timepoint == 'TF') %>% #we want the final volume
  mutate(ZFMx = (VFM/volume_ml)*ZFM+(VFWC/volume_ml)*ZFWC, #Control final corrected for sample volume removed
         ZavWC = (Z0-ZFWC)/log(Z0/ZFWC), #Average concentration chla in the water column
         ZavM = (Z0-ZFM)/log(Z0/ZFM)) %>%
  # create a label column for easier handling
  mutate(label = paste(Culture, GROUP, NPratio, sep = "_")) %>%
  mutate(category = 'grn') %>%
  select(-c(Timepoint, VFM, VFWC))

## Process bluegreen readings
blugrn_mix_summary <- mix_data %>%
  filter(category == 'blugrn') %>% 
  select(Date, Timepoint, REP, Culture, GROUP, NPratio, NOTE, mean_chla) %>%
  # check for duplicates
  
  group_by(Date, Timepoint, REP, Culture, GROUP, NPratio, NOTE) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_chla) %>%
  mutate(
  # let Z0 be the initial concentration at T0 or the second lowest timepoint
    Z0 = ifelse(is.na(T0), T1, T0),
    #Final control measurement before mixing
    ZFWC = ifelse(is.na(T5), T4, T5), 
    #Final control measurement after mixing
    ZFM = TF) %>%
  select(-c(T0, T1, T2, T3, T4, T5, TF)) %>%
  left_join(mix_data %>% 
              filter(category == 'blugrn') %>%
              select(-c(mean_chla,mean_time,category)), 
            by = c('Date', 'REP', 'Culture', 'GROUP', 'NPratio', 'NOTE'), keep = FALSE) %>%
  #Control final corrected for sample volume removed
  filter(Timepoint == 'TF') %>% #we want the final volume
  mutate(ZFMx = (VFM/volume_ml)*ZFM+(VFWC/volume_ml)*ZFWC, #Control final corrected for sample volume removed
         ZavWC = (Z0-ZFWC)/log(Z0/ZFWC), #Average concentration chla in the water column
         ZavM = (Z0-ZFM)/log(Z0/ZFM)) %>%
  # create a label column for easier handling
  mutate(label = paste(Culture, GROUP, NPratio, sep = "_")) %>%
  mutate(category = 'blugrn') %>%
  select(-c(Timepoint, VFM, VFWC))

total_chla_mix_summary <- mix_data %>%
  filter(category == 'total_chla') %>% 
  select(Date, Timepoint, REP, Culture, GROUP, NPratio, NOTE, mean_chla) %>%
  # check for duplicates
  
  group_by(Date, Timepoint, REP, Culture, GROUP, NPratio, NOTE) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_chla) %>%
  mutate(
  # let Z0 be the initial concentration at T0 or the second lowest timepoint
    Z0 = ifelse(is.na(T0), T1, T0),
    #Final control measurement before mixing
    ZFWC = ifelse(is.na(T5), T4, T5), 
    #Final control measurement after mixing
    ZFM = TF) %>%
  select(-c(T0, T1, T2, T3, T4, T5, TF)) %>%
  left_join(mix_data %>% 
              filter(category == 'total_chla') %>%
              select(-c(mean_chla,mean_time,category)), 
            by = c('Date', 'REP', 'Culture', 'GROUP', 'NPratio', 'NOTE'), keep = FALSE) %>%
  #Control final corrected for sample volume removed
  filter(Timepoint == 'TF') %>% #we want the final volume
  mutate(ZFMx = (VFM/volume_ml)*ZFM+(VFWC/volume_ml)*ZFWC, #Control final corrected for sample volume removed
         ZavWC = (Z0-ZFWC)/log(Z0/ZFWC), #Average concentration chla in the water column
         ZavM = (Z0-ZFM)/log(Z0/ZFM)) %>%
  # create a label column for easier handling
  mutate(label = paste(Culture, GROUP, NPratio, sep = "_")) %>%
  mutate(category = 'total_chla') %>%
  select(-c(Timepoint, VFM, VFWC))

mix_summary <- bind_rows(grn_mix_summary, blugrn_mix_summary,total_chla_mix_summary) %>%
  # join AFDW data
  left_join(feeding_data %>%
              filter(Timepoint == 'TF') %>%
              select(Date, Culture, GROUP, REP, NPratio, 
                     NOTE, category, hours,total_AFDW, total_C), 
            by = c('Date', 'Culture', 'GROUP', 'REP',
                   'NPratio', 'NOTE', 'category'), 
            keep = FALSE)

# Visualize the summary
mix_plot <- ggplot(mix_summary, aes(x = NPratio, y = ZFMx)) +
  geom_boxplot(aes(fill = category), position = position_dodge(width = 0.5), outlier.shape = NA, alpha = 0.3) +
   scale_fill_manual(values = c("blugrn" = "cadetblue",
                                "grn" = "yellowgreen",
                                "total_chla" = "plum"),
                                labels = c("blugrn" = "*M. aeruginosa*",
                                           "grn" = "*C. oblonga*",
                                           "total_chla" = "Total phytoplanktons"), 'category') +
  geom_point(aes(fill = Date),size = 1, position = position_dodge(width = 0.1)) +
  facet_wrap(~Culture, nrow = 1, labeller = labeller(Culture = c("LE19-195" = "Mix 1: *M. aeruginosa* LE19-195",
                                                            "LE19-131" = "Mix 2: *M. aeruginosa* LE19-131",
                                                            "LE19-114" = "Mix 3: *M. aeruginosa* LE19-114"))) +
  labs(title = 'Volume-corrected TF Chl a Concentration in Experimental Beakers',
    y = expression("Chl a ("*mu*"g/L)"),
    x = "Experimental Beakers (Culture_NPratio)") +
  theme_bw()+
  theme(legend.position = 'right') +
  rotate_x_text(angle = 60)
mix_plot
```

### Calculate Clearance and Ingestion Rates with correction for growth and death in controls

```{r}
#Combine the experimental and control data to calculate feeding and clearance rates
combined_data <-merge(mix_summary, control_summary,by= c('Date','Culture', 'NPratio','NOTE','category')) %>% #Merge control data with exp data.
  mutate(FR = (volume_ml/(total_AFDW*hours))*(log(Z0)-log(ZFWC)+log(CFWC)-log(C0)), #Clearance rate on amount removed, includes correction
         FI = (volume_ml/(total_AFDW*hours))*(log(Z0)-log(ZFMx)+log(CFMx)-log(C0)), #Clearance rate on amount ingested, includes correction
         CR = ZavWC*FR/1000,#Capture rate
         IR = ZavM*FI/1000, #Ingestion rate
         )

#Change all Inf, NA, NaN values to 0, as these represent treatments where either bluegreen or green were not present. E.g. C. oblonga controls
combined_data[sapply(combined_data, is.infinite)] <- NA
combined_data[sapply(combined_data, is.nan)] <- NA

write.table(combined_data, "output_tables/NP_QMF_25.out.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

(From Vanderploeg 2001) C is the mean Chl a concentration in the water column at the end of the experiment in control containers, Z is the Chl a concentration in the water column of an experimental container at the end of the experiment, V is the volume of water in the container during the experiment, n is the number of mussels, and t is the duration of the experiment.

## Read data and set parameters

```{r}
combined_data <- read.table("output_tables/NP_QMF_25.out.txt", header = TRUE, sep = "\t") %>%
  # make NPratio a factor
  mutate(NPratio = factor(NPratio, levels = c( '40', '20', '5', '80', 'LOW20'))) %>%
  filter(NPratio %in% c('5', '20', '80'))
# Define colors and labels for strains
my_colors <- c("LE19-195" = "lightblue4",
                                "LE19-131" = "darkorange2",
                                "LE19-114" = "lightpink4",
                               "CHLAMY" = "darkgreen")
my_labels <- c(
  "LE19-195" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-195") ~ bold("(Selective strain)")),
  #"LE19-131" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-131") ~ bold("(Selective strain)")),
  "LE19-114" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-114") ~ bold("(Inhibitory strain)")),
  "CHLAMY"   = expression(bolditalic("Chlamydomonas oblonga") ~ bold("(Preferred food)")))
```

## Visualizing CLearance Rate Result

```{r} 
# FI but only NP5,20 and 80
FI_subset_data <- combined_data %>% 
         filter(Culture %in% c("LE19-114","LE19-131","LE19-195")) %>%
         filter(category %in% c('grn','blugrn')) %>%
         filter(NPratio %in% c('5', '20', '80')) %>%
         mutate(Culture = factor(Culture, levels = c( "LE19-131", 'LE19-195','LE19-114'))) %>%
         mutate(NPratio = factor(NPratio, levels = c( '5', '20','80'))) %>%
         mutate(strain_resolved_cat = case_when(
           Culture == "LE19-114" & category == "blugrn" ~ "LE19-114",
           Culture == "LE19-131" & category == "blugrn" ~ "LE19-131",
           Culture == "LE19-195" & category == "blugrn" ~ "LE19-195",
           TRUE ~ category
         ))
ggplot(FI_subset_data,
       aes(x = NPratio, y = FI)) +
  geom_boxplot(aes(fill = strain_resolved_cat), position = position_dodge(width = 1)) +
  # add jitter and separate points by category
  geom_point(aes(fill = strain_resolved_cat),size = 1.5, alpha = 0.5, 
             position = position_jitterdodge(jitter.width = 0.2, dodge.width = 1)) +
  facet_wrap(~Culture, nrow = 3, labeller = labeller(Culture = my_labels)) +
  scale_fill_manual(name = 'Cultures',
                    values = c("LE19-131" = "grey50","LE19-195" = "goldenrod2",
                               "LE19-114" = "lightpink4","grn" = "snow"),
                    labels = c("grn" = "Preferred prey",
                               "LE19-131" = "Unpalatable strain (group1)",
                               "LE19-195" = "Unpalatable strain (group 3)",
                               "LE19-114" = "Inhibitory strain (group 1)"
                               )) +
  scale_y_continuous(limits = c(-10, NA)) +
  labs(
    y = expression(bold("Amount ingested by mussel ("*bold(mL)*"/mg/"*bold(hr)*")")),
    x = "N:P ratios") +
  theme_pubclean()+
  theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(size=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=12,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold"),
        axis.title = element_text(size=15,color='black',face = "bold"),
        strip.text = element_text(size = 12, color = 'black',face = "bold"),
        strip.background = element_rect(fill = "grey100", 
                                        color = NA, linewidth = 1),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=15,color='black',hjust=0, face = "bold"),
        legend.text = element_text(size=12,color='black',hjust=0),
        legend.position = 'right',
        panel.grid.major = element_blank()) +
  rotate_x_text(angle = 60)
ggsave("figures/NP_QMF_25_FI_plot_subset.png", width = 6, height = 8)
```

Plot the raw data of those negative feeding rates

```{r}
neg_FI_data <- combined_data %>%
  filter(Culture != 'CHLAMY') %>%
  filter(NPratio %in% c('5', '20', '80')) %>%
  filter(category != 'total_chla') %>%
  mutate(NPratio = factor(NPratio, levels = c( '5', '20','80'))) %>%
  mutate(Culture = factor(Culture, levels = c( "LE19-131", 'LE19-195','LE19-114'))) %>%
  mutate(neg_FI = ifelse(FI < 0, TRUE, FALSE)) %>%
  mutate(quagga_log_diff = log(Z0) - log(ZFMx),
         control_log_diff = log(C0) - log(CFMx)) %>%
  select(Date, Culture, REP.x, NPratio, category, quagga_log_diff, control_log_diff, neg_FI) %>%
  pivot_longer(c(quagga_log_diff, control_log_diff), 
               names_pattern = "(.*)_log_diff",
               names_to = "Treatment", values_to = "log_diff") 

ggplot(neg_FI_data, aes(x = NPratio, y = log_diff)) +
  geom_boxplot(aes(fill = Treatment), position = position_dodge(width = 0.5), outlier.shape = NA, alpha = 0.5) +
  geom_point(aes(group = Treatment, color = neg_FI,shape = neg_FI),size = 2, alpha = 0.5, position = position_dodge(width = 0.5)) +
  facet_grid(rows = vars(Culture), cols = vars(category), labeller = labeller(Culture = my_labels, category = c("blugrn" = "Microcystis", "grn" = "Preferred prey"))) +
  scale_fill_manual(values = c("quagga" = "plum4","control" = "lightgreen"),
                                labels = c("quagga" = "Quagga mussel",
                                           "control" = "Control"), 'Treatment') +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"),
                     labels = c("TRUE" = "Negative value",
                                "FALSE" = "Positive value"), c('neg_FI' = 'Amount ingeted by mussels')) +
  scale_shape_manual(values = c("TRUE" = 4, "FALSE" = 16),
                     labels = c("TRUE" = "Negative value",
                                "FALSE" = "Positive value"), c('neg_FI' = 'Amount ingeted by mussels')) +
  labs(title = NULL,
    y = "Log difference in chl-A concentration (mg; initial-final)",
    x = "N:P ratios") +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(size=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=15,color='black',face = "bold"),
        strip.text = element_text(size = 12, color = 'black',face = "bold"),
        strip.background = element_rect(fill = "grey100", 
                                        color = NA, linewidth = 1),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=15,color='black',hjust=0, face = "bold"),
        legend.text = element_text(size=12,color='black',hjust=0),
        legend.position = 'right',
        panel.grid.major = element_blank()) +
  rotate_x_text(angle = 60)

ggsave("figures/supplFig_on_negative_FI_plot.png", width = 8, height = 8)
```



```{r}
# Visualize Total Ingestion by Culture and NPratio
FI_traits <- combined_data %>%
  # There is no blugrns in the CHLAMY control. The FI value only indicate noise, so set it to zero.
  mutate(FI = ifelse(Culture == 'CHLAMY' & category == 'blugrn', 0, FI)) %>%
  select(Date, Culture, REP.x, NPratio, category, FI) %>%
  pivot_wider(names_from = category, values_from = FI) %>%
  drop_na() %>%
  mutate(total_FI = total_chla,
         blugrn_grn_ratio = blugrn/grn,
         selective_proportion = (grn-blugrn)/(grn + blugrn))
total_FI_plot <- ggplot(FI_traits, aes(x = NPratio, y = total_FI)) +
  geom_boxplot(aes(fill = NPratio), position = position_dodge(width = 1), outlier.shape = NA, alpha = 0.3) +
  geom_point(aes(color = Date),size = 3, position = position_dodge(width = 0.5)) +
  facet_wrap(~Culture, nrow = 1, labeller = labeller(Culture = c("LE19-195" = "Mix 1: *M. aeruginosa* LE19-195",
                                                            "LE19-131" = "Mix 2: *M. aeruginosa* LE19-131",
                                                            "LE19-114" = "Mix3: *M. aeruginosa* LE19-114"))) +
  scale_fill_manual(values = c("5" = "lightblue","20" = "lightgreen", "80" = "lightpink"),
                                labels = c("5" = "N:P 5:1",
                                           "20" = "N:P 20:1",
                                           "80" = "N:P 80:1"), 'NPratio') +
  labs(title = 'Total Ingestion Rate in Experimental Beakers',
    y = expression("Total Ingestion Rate ("*mL*"/mg AFDW/"*hr*")"),
    x = "Experimental Beakers (Culture_NPratio)") +
  theme_bw()+
  theme(legend.position = 'right') +
  rotate_x_text(angle = 60)

blugrn_grn_ratio_plot <- ggplot(FI_traits, aes(x = NPratio, y = blugrn_grn_ratio)) +
  geom_boxplot(aes(fill = NPratio), position = position_dodge(width = 0.5), outlier.shape = NA, alpha = 0.3) +
  geom_point(aes(color = Date),size = 1.5, alpha = 0.5, position = position_dodge(width = 0.5)) +
  facet_wrap(~Culture, nrow = 1, labeller = labeller(Culture = c("LE19-195" = "Mix 1: *M. aeruginosa* LE19-195",
                                                            "LE19-131" = "Mix 2: *M. aeruginosa* LE19-131", 
                                                            "LE19-114" = "Mix3: *M. aeruginosa* LE19-114"))) +
  scale_fill_manual(values = c("5" = "lightblue","20" = "lightgreen", "80" = "lightpink"),
                                labels = c("5" = "N:P 5:1",
                                           "20" = "N:P 20:1",
                                           "80" = "N:P 80:1"), 'NPratio') +
  scale_y_continuous(limits = c(-1, 2)) +
  labs(title = 'Bluegreen to Green Ingestion Ratio in Experimental Beakers',
    y = expression("Bluegreen to Green Ingestion Ratio"),
    x = "Experimental Beakers (Culture_NPratio)") +
  theme_bw()+
  theme(legend.position = 'right') +
  rotate_x_text(angle = 60)

total_FI_plot;blugrn_grn_ratio_plot
```


# Modeling Mussel Selectivity and paramterizing with Microcystis strain and NPratio

## Test feeding difference on the preferred food across NPratio levels

We have to first deal with the baseline model of feeding as a function of NPratio because feeding seems to increase with NP ratio in the preferred food control.

```{r}
# Get CHLAMY data
chlamy_FI <- combined_data %>%
  filter(Culture == 'CHLAMY' & category == "grn") %>%
  select(Date, Culture, NPratio, category, FI) %>%
  drop_na() 
# visualize chlamy feeding by NPratio
chlamy_FI_plot <- ggplot(chlamy_FI, aes(x = NPratio, y = FI)) +
  geom_point(size = 3, alpha = 0.7, color = 'darkgreen') +
  geom_smooth(method = 'lm', aes(group = Culture),
              alpha = 0.2,se = TRUE, linetype = 'dashed', color = 'darkgreen', fill = 'lightgreen') +
  #ggrepel::geom_text_repel(aes(label = Date), size = 3, color = 'black', max.overlaps = 10) +
  labs(title = 'Chlamydomonas oblonga Feeading Rate by N:P Ratio',
       x = 'N:P Ratio',
       y = expression("Clearnace Rate on Amount Ingested ("*mL*"/mg AFDW/"*hr*")")) +
  theme_bw()
chlamy_FI_plot
```

Fit baseline model

```{r}
library(lmerTest)
# Fit linear model to chlamy feeding data
chlamy_FI_lm1 <- lmerTest::lmer(FI ~ NPratio+(1|Date), data = chlamy_FI)
chlamy_FI_lm2 <- lm(FI ~ NPratio, data = chlamy_FI)
chlamy_FI_lm3 <- lmerTest::lmer(FI ~ (1|Date), data = chlamy_FI)
# Compare models
AIC(chlamy_FI_lm1, chlamy_FI_lm2, chlamy_FI_lm3)
# Summarize the best model
summary(chlamy_FI_lm1)
# ANOVA to test N:P ratio effect
car::Anova(chlamy_FI_lm1, type = 3)
```

There is no main effect of NPratio on feeding rate in Chlamydomonas oblonga controls. We will use the intercept of the model as the baseline feeding rate for mussels.

```{r}
baseline_FI <- fixef(chlamy_FI_lm1)[1]
baseline_FI
```

## Visualizing selectivity and feeding inhibition results

Let's model mussel selectivity as a function of Microcystis strain type and NPratio. We will calculate selectivity as the ratio of bluegreen ingestion rate to green ingestion rate (BG:G FI ratio).

```{r}
# Load data 
combined_data <- read.table("output_tables/NP_QMF_25.out.txt", header = TRUE, sep = "\t")

FI_traits <- combined_data %>%
  #filter(Culture != 'LE19-131') %>%
  # The negative FI values for blugrns indicate no feeding on Microcystis when Chlamydomonas is present, so it should be set to zero.
  #mutate(FI = ifelse(FI < 0, 0+0.00001, FI)) %>%
  # There is no blugrns in the CHLAMY control. The FI value only indicate noise, so set it to zero.
  mutate(FI = ifelse(Culture == 'CHLAMY' & category == 'blugrn', 0+0.00001, FI)) %>%
  select(Date, Culture, REP.x, NPratio, category, FI) %>%
  pivot_wider(names_from = category, values_from = FI) %>%
  drop_na() %>%
  mutate(total_FI = total_chla,
         blugrn_grn_ratio = blugrn/grn,
         selective_proportion = (grn-blugrn)/(grn + blugrn))

# Identify outliers in LE19-114 NP 80. The outlier is a result of a very low green feeding rate, which is because the mussels cleared all of the green alga.
outlier_indices <- which(FI_traits$Culture == 'LE19-114' & FI_traits$NPratio == '80' & abs(FI_traits$blugrn_grn_ratio) > 2)

  
  
FI_traits_clasic <- FI_traits %>%
  slice(-outlier_indices) %>%
  filter(NPratio %in% c('5','20','80')) %>%
  # make NPratio a factor
  mutate(NPratio = factor(NPratio, levels = c( '5', '20','80')))
# Define colors and labels for strains
my_colors <- c("LE19-195" = "lightblue4",
                                "LE19-131" = "darkorange2",
                                "LE19-114" = "lightpink4",
                               "CHLAMY" = "darkgreen")
my_labels <- c(
  "LE19-195" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-195") ~ bold("(Selective strain)")),
  #"LE19-131" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-131") ~ bold("(Selective strain)")),
  "LE19-114" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-114") ~ bold("(Inhibitory strain)")),
  "CHLAMY"   = expression(bolditalic("Chlamydomonas oblonga") ~ bold("(Preferred food)")))

# scatter plot of unpalatability per sttrain
unpalatability_plot <- ggplot(FI_traits_clasic, aes(x = NPratio, y = blugrn_grn_ratio, color = Culture)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(aes(group = Culture,color = Culture, fill = Culture),method = 'lm',
              formula = y ~ poly(x, 2),
              alpha = 0.1,se = TRUE, linetype = 'dashed') +
  scale_fill_manual(name = 'Culture',
                     values = my_colors,
                     labels = my_labels) +
  scale_color_manual(name = 'Culture',
                     values = my_colors,
                     labels = my_labels) +
  labs(title = 'Unpalatability',
       x = 'N:P Ratio',
       y = expression(bold('Feeding Ratio ('*bolditalic(M.aeruginosa)~': Preferred food)'))) +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(size=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold"),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.position = 'right',
          panel.grid.major = element_blank())
# scatterplot of inhibition per strain
inhibition_plot <- ggplot(FI_traits_clasic, aes(x = NPratio, y = total_FI, color = Culture)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(aes(group = Culture,color = Culture, fill = Culture),method = 'lm',
              formula = y ~ poly(x, 2),
              alpha = 0.1,se = TRUE, linetype = 'dashed') +
  scale_fill_manual(name = 'Culture',
                     values = my_colors,
                     labels = my_labels) +
  scale_color_manual(name = 'Culture',
                     values = my_colors,
                     labels = my_labels) +
  labs(title = 'Inhibition',
       x = 'N:P Ratio',
       y = 'Total amount of phytoplanktons ingested (mL/mg AFDW/hr)') +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(size=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold"),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black', face = "bold"),
          legend.position = 'right',
          panel.grid.major = element_blank())
#ggsave("NP_QMF_25_selectivity_plot_classic.png", selectivity_plot, width = 8, height = 6)
#ggsave("NP_QMF_25_total_ingestion_plot_classic.png", total_ingestion_plot, width = 8, height = 6)
unpalatability_plot; inhibition_plot
```


```{r}
select_FI_traits <- FI_traits %>%
  slice(-outlier_indices) %>%
  filter(Culture != 'CHLAMY') %>%
  filter(NPratio %in% c('5','20','80')) %>%
  mutate(Culture = as.factor(Culture)) %>%
  mutate(NPratio = factor(NPratio, levels = c('5', '20', '80'))) %>%
  mutate(NP_ratio_numeric = case_when(
           NPratio == '5' ~ 5,
           NPratio == '20' ~ 20,
           NPratio == '80' ~ 80
  ))
```


```{r}
# Fit linear mixed model
# We removed the intercept term based on the biological constraint that blugrn_grn_ratio cannot exist when NPratio = 0
unpalatabilty_lm1 <- lmerTest::lmer(blugrn_grn_ratio ~ Culture * NPratio -1 + (1|Date), data = select_FI_traits)
unpalatabilty_lm2 <- lmerTest::lmer(blugrn_grn_ratio ~ Culture * NPratio + (1|Date), data = select_FI_traits)
unpalatabilty_lm3 <- lmerTest::lmer(blugrn_grn_ratio ~ Culture + NPratio -1 + (1|Date), data = select_FI_traits)
unpalatabilty_lm4 <- lmerTest::lmer(blugrn_grn_ratio ~ Culture -1 + (1|Date), data = select_FI_traits)
unpalatabilty_lm5 <- lm(blugrn_grn_ratio ~ Culture * NPratio-1, data = select_FI_traits)
# Compare models
AIC(unpalatabilty_lm1, unpalatabilty_lm2, unpalatabilty_lm3, unpalatabilty_lm4, unpalatabilty_lm5)
# Summarize the best model
summary(unpalatabilty_lm1)
```

Export summary table of the best model

```{r}
# Export summary table of the unpalatability_lm1
unpal_fixed_effects_summary <- tidy(unpalatabilty_lm1, effects = "fixed") %>%
  mutate_if(is.numeric, round, digits = 3)
unpal_random_effects_summary <- tidy(unpalatabilty_lm1, effects = "ran_pars") %>%
  mutate_if(is.numeric, round, digits = 3)

write.xlsx(list("Fixed Effects" = unpal_fixed_effects_summary,
                "Random Effects" = unpal_random_effects_summary),
           file = "output_tables/unpalatability_lm1_summary.xlsx")
```

```{r}
# ANOVA to test main and interaction effects
unpal_anova_result <- car::Anova(unpalatabilty_lm1, type = "III", test.statistic = "F")

unpal_anova_df <- as.data.frame(unpal_anova_result) %>%
  rownames_to_column(var = "Term") %>%
  mutate(F = round(F, 2),
        Df.res = round(Df.res, 2),
         `Pr(>F)` = case_when(
           `Pr(>F)` < 0.001 ~ "<0.001",
           TRUE ~ sprintf("%.3f", `Pr(>F)`)))

unpal_table<- flextable::flextable(unpal_anova_df) %>%
  flextable::set_header_labels(Term = "Term", Df = "Df", `Df.res` = "Residual DF", F = "F value", `Pr(>F)` = "p-value") %>%
  flextable::italic(j = "F", part = "header") %>%
  flextable::italic(j = "Pr(>F)", part = "header") %>%
  flextable::theme_vanilla() %>%
  flextable::bold(i = ~ `Pr(>F)` < 0.05, j = "Pr(>F)", bold = TRUE) %>%
  flextable::save_as_docx(path = "output_tables/unpalatability_lm1_anova.docx")
```

```{r}
# assumptions
par(mfrow=c(2,2))
plot(unpalatabilty_lm1, xlab = "Fitted values", ylab = "Residuals", main = "Model for unpalatability")
par(mfrow=c(1,1))

## normality
qqnorm(resid(unpalatabilty_lm1), main = "Normal Q-Q Plot \n (unpalatability model)")
qqline(resid(unpalatabilty_lm1))
shapiro.test(resid(unpalatabilty_lm1))

## homogeneity of variance
car::leveneTest(resid(unpalatabilty_lm1) ~ select_FI_traits$Culture)
car::leveneTest(resid(unpalatabilty_lm1) ~ select_FI_traits$NPratio)

```

```{r}
# post hoc test: multiple comparisons over NPratio in cultures
unpalatabilty_emm <- emmeans::emmeans(unpalatabilty_lm1, pairwise ~ NPratio | Culture, type = 'response', adjust = 'tukey')

unpal_contrast_table <- as.data.frame(unpalatabilty_emm$contrasts) %>%
  mutate(contrast = gsub("NPratio = ", "", contrast),
         estimate = round(estimate, 3),
         SE = round(SE, 3),
         df = round(df, 2),
         t.ratio = round(t.ratio, 2),
         p.value = case_when(
           p.value < 0.001 ~ "<0.001",
           TRUE ~ sprintf("%.3f", p.value)
         )) %>%
  select(Culture, contrast, estimate, SE, df, t.ratio, p.value)

flextable::flextable(unpal_contrast_table) %>%
  flextable::set_header_labels(Culture = "Culture", contrast = "Contrast", estimate = "Estimate", SE = "SE", df = "DF", t.ratio = "t-ratio", p.value = "p-value") %>%
  flextable::italic(j = c("t.ratio", "p.value"), part = "header") %>%
  flextable::theme_vanilla() %>%
  flextable::bold(i = ~ p.value < 0.05, j = "p.value", bold = TRUE) %>%
  flextable::save_as_docx(path = "output_tables/unpalatability_lm1_posthoc.docx")
```

(5vs20: estimate = -0.54, 95%CI[-0.71,-0.37], p-value < 0.01; 

Inhibition

```{r}
# We removed the intercept term based on the biological constraint that blugrn_grn_ratio cannot exist when NPratio = 0
inhibition_lm1 <- lmerTest::lmer(total_FI ~ Culture * NPratio -1 + (1|Date), data = select_FI_traits)
inhibition_lm2 <- lmerTest::lmer(total_FI ~ Culture * NPratio + (1|Date), data = select_FI_traits)
inhibition_lm3 <- lmerTest::lmer(total_FI ~ Culture + NPratio -1 + (1|Date), data = select_FI_traits)
inhibition_lm4 <- lmerTest::lmer(total_FI ~ Culture -1 + (1|Date), data = select_FI_traits)
inhibition_lm5 <- lm(total_FI ~ Culture * NPratio -1, data = select_FI_traits)
# Compare models
AIC(inhibition_lm1, inhibition_lm2, inhibition_lm3, inhibition_lm4, inhibition_lm5)
# Summarize the best model
summary(inhibition_lm1)
```

Export summary table of the best model

```{r}
inhib_fixed_effects_summary <- tidy(inhibition_lm1, effects = "fixed") %>%
  mutate_if(is.numeric, round, digits = 3)
inhib_random_effects_summary <- tidy(inhibition_lm1, effects = "ran_pars") %>%
  mutate_if(is.numeric, round, digits = 3)

write.xlsx(list("Fixed Effects" = inhib_fixed_effects_summary,
                "Random Effects" = inhib_random_effects_summary),
           file = "output_tables/inhibition_lm1_summary.xlsx")
```

```{r}
# ANOVA to test main and interaction effects
inhib_anova_result <- car::Anova(inhibition_lm1, type = "III", test.statistic = "F")

inhib_anova_df <- as.data.frame(inhib_anova_result) %>%
  rownames_to_column(var = "Term") %>%
  mutate(F = round(F, 2),
        Df.res = round(Df.res, 2),
         `Pr(>F)` = case_when(
           `Pr(>F)` < 0.001 ~ "<0.001",
           TRUE ~ sprintf("%.3f", `Pr(>F)`)))

inhib_table<- flextable::flextable(inhib_anova_df) %>%
  flextable::set_header_labels(Term = "Term", Df = "Df", `Df.res` = "Residual DF", F = "F value", `Pr(>F)` = "p-value") %>%
  flextable::italic(j = "F", part = "header") %>%
  flextable::italic(j = "Pr(>F)", part = "header") %>%
  flextable::theme_vanilla() %>%
  flextable::bold(i = ~ `Pr(>F)` < 0.05, j = "Pr(>F)", bold = TRUE) %>%
  flextable::save_as_docx(path = "output_tables/inhibition_lm1_anova.docx")
inhib_anova_df
```

```{r}
# assumptions
par(mfrow=c(2,2))
plot(inhibition_lm1, xlab = "Fitted values", ylab = "Residuals", main = "Model for inhibition")
par(mfrow=c(1,1))

## normality
qqnorm(resid(inhibition_lm1), main = "Normal Q-Q Plot \n (inhibition model)")
qqline(resid(inhibition_lm1))
shapiro.test(resid(inhibition_lm1))

## homogeneity of variance
car::leveneTest(resid(inhibition_lm1) ~ select_FI_traits$Culture)
car::leveneTest(resid(inhibition_lm1) ~ select_FI_traits$NPratio)

```


```{r}
# post hoc test: multiple comparisons over NPratio in cultures
inhib_emm <- emmeans::emmeans(inhibition_lm1, pairwise ~ NPratio | Culture,  adjust = 'tukey')
inhib_contrast_table <- as.data.frame(inhib_emm$contrasts) %>%
  mutate(contrast = gsub("NPratio = ", "", contrast),
         estimate = round(estimate, 3),
         SE = round(SE, 3),
         df = round(df, 2),
         t.ratio = round(t.ratio, 2),
         p.value = case_when(
           p.value < 0.001 ~ "<0.001",
           TRUE ~ sprintf("%.3f", p.value)
         )) %>%
  select(Culture, contrast, estimate, SE, df, t.ratio, p.value)

flextable::flextable(inhib_contrast_table) %>%
  flextable::set_header_labels(Culture = "Culture", contrast = "Contrast", estimate = "Estimate", SE = "SE", df = "DF", t.ratio = "t-ratio", p.value = "p-value") %>%
  flextable::italic(j = c("t.ratio", "p.value"), part = "header") %>%
  flextable::theme_vanilla() %>%
  flextable::bold(i = ~ p.value < 0.05, j = "p.value", bold = TRUE) %>%
  flextable::save_as_docx(path = "output_tables/inhibition_lm1_posthoc.docx")
```

```{r}
inhib_19114_lm <- lmerTest::lmer(total_FI ~ NPratio -1 + (1|Date), data = select_FI_traits %>% filter(Culture == 'LE19-114'))
inhib_19195_lm <- lmerTest::lmer(total_FI ~ NPratio -1 + (1|Date), data = select_FI_traits %>% filter(Culture == 'LE19-195'))
inhib_19131_lm <- lmerTest::lmer(total_FI ~ NPratio -1 + (1|Date), data = select_FI_traits %>% filter(Culture == 'LE19-131'))

car::Anova(inhib_19114_lm, type = "III", test.statistic = "F")
car::Anova(inhib_19195_lm, type = "III", test.statistic = "F")
car::Anova(inhib_19131_lm, type = "III", test.statistic = "F")

```


```{r}
# assumptions
par(mfrow=c(2,2))
plot(inhib_19195_lm)
par(mfrow=c(1,1))

## normality
qqnorm(resid(inhib_19195_lm))
qqline(resid(inhib_19195_lm))
shapiro.test(resid(inhib_19195_lm))

## homogeneity of variance
car::leveneTest(resid(inhib_19195_lm) ~ select_FI_traits %>% filter(Culture == "LE19-195") %>% pull(NPratio))
```

```{r}
# post hoc test: multiple comparisons over NPratio in cultures
inhib_195_emm <- emmeans::emmeans(inhib_19195_lm, pairwise ~ NPratio,  adjust = 'tukey')
inhib_195_emm$contrasts
```



```{r}
# assumptions
par(mfrow=c(2,2))
plot(inhib_19131_lm)
par(mfrow=c(1,1))

## normality
qqnorm(resid(inhib_19131_lm))
qqline(resid(inhib_19131_lm))
shapiro.test(resid(inhib_19131_lm))

## homogeneity of variance
car::leveneTest(resid(inhib_19131_lm) ~ select_FI_traits %>% filter(Culture == "LE19-131") %>% pull(NPratio))
```

```{r}
# post hoc test: multiple comparisons over NPratio in cultures
inhib_131_emm <- emmeans::emmeans(inhib_19131_lm, pairwise ~ NPratio,  adjust = 'tukey')
inhib_131_emm$contrasts
```

Plot total feeding rates to selectivity

```{r}
library(ggnewscale)
library(ggrepel)
plot_traits_NP <- FI_traits %>% 
  slice(-outlier_indices) %>%
  filter(Culture != 'CHLAMY') %>%
  filter(!NPratio %in% c('40','LOW20')) %>%
  mutate(NP_ratio_numeric = case_when(
           NPratio == '5' ~ 5,
           NPratio == '20' ~ 20,
           NPratio == '80' ~ 80
           #NPratio == '40' ~ 40
           #NPratio == 'LOW20' ~ 20
  )) %>%
  group_by(Culture, NPratio,NP_ratio_numeric) %>%
  summarise(mean_palatability = mean(blugrn_grn_ratio),
            mean_total_FI = mean(total_FI),
            se_palatability = sd(blugrn_grn_ratio)/sqrt(n()),
            se_total_FI = sd(total_FI)/sqrt(n())) %>%
  mutate(NPratio = as.factor(NPratio)) %>%
  mutate(np_labels = case_when(
    NPratio == "5" ~ "N:P=5",
    NPratio == "20" ~ "N:P=20",
    NPratio == "LOW20" ~ "N:P=20 (Low absolute concentration)",
    NPratio == "40" ~ "N:P=40 (P-limitation)",
    NPratio == "80" ~ "N:P=80",
    TRUE ~ NPratio
  ))
# my shapes
my_shapes <- c("5" = 15,"20" = 16, "80" = 17, "LOW20" = 18, "40" = 8)
my_np_labels <- c("5" = "N:P=5:1 (N-limitation)",
                   "20" = "N:P 20:1",
                  "LOW20" = "N:P 20:1 (Low absolute concentration)",
                  "40" = "N:P 40:1 (P-limitation)",
                   "80" = "N:P 80:1 (P-limitation)")
# Define colors and labels for strains
my_colors <- c("LE19-195" = "goldenrod",
                                "LE19-131" = "grey40",
                                "LE19-114" = "lightpink4",
                               "CHLAMY" = "darkgreen")
my_labels <- c(
  "LE19-195" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-195") ~ bold("(Unpalatable strain)")),
  "LE19-131" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-131") ~ bold("(Unpalatable strain)")),
  "LE19-114" = expression(bolditalic("M.aeruginosa") ~ bold("LE19-114") ~ bold("(Inhibitory strain)")),
  "CHLAMY"   = expression(bolditalic("Chlamydomonas oblonga") ~ bold("(Preferred prey)")))


## traits moving along the nutrient landscape
ggplot(plot_traits_NP, aes(x = mean_palatability, y = mean_total_FI)) +
  geom_point(aes(shape = NPratio, colour = Culture),size = 5, alpha = 1, stroke = 1.2) +
  ggrepel::geom_text_repel(aes(label = NPratio), 
                           max.overlaps = 30,hjust = 0, vjust = 0,
                           size = 4, alpha = 1,fontface = 'bold') +
  geom_linerange(aes(ymin = mean_total_FI - se_total_FI, ymax = mean_total_FI + se_total_FI, color = Culture), linewidth = 0.5, alpha = 0.7) +
  geom_linerange(aes(xmin = mean_palatability - se_palatability, xmax = mean_palatability + se_palatability, color = Culture), linewidth = 0.5, alpha = 0.7) +
  scale_color_manual(name = 'Culture',
                     values = my_colors,
                     labels = my_labels) +
  scale_shape_manual(name = "Nutrient regime", 
                     values = my_shapes,
                     labels = my_np_labels) +
  labs(title = NULL,
       x = expression(bold('Feeding Selectivity ('*bolditalic(M.aeruginosa)~': Preferred food)')),
       y = 'Total amount of phytoplankton ingested (mL/mg AFDW/hr)') +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(size=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=12,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold"),
          axis.title = element_text(size=12,color='black',face = "bold"),
          #panel.background = element_rect(fill = "snow"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.position = 'right',
          panel.grid.major = element_blank())
ggsave("figures/NP_QMF_25_traits_NP_landscape.png", width = 7.5, height = 4)

```



