---
title: "figS9_effect_sizes"
author: "Yu-Cheng Lin"
date: "2026-02-08"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(circlize)
library(car)
library(broom)
source('scripts/parse_functions.R')
```


# Defense trait

Load defense trait data

```{r}
# Trait data summary from dreissinid mussel feeding experiments
raw_defense_trait <- read.table('inputs/LE_QMF_23_IDM_trait_summary.out.txt', header = T, sep = '\t')
defense_trait <- raw_defense_trait%>%
  mutate(Strain = str_extract(Culture, '^[^.]+')) %>%
  mutate(Strain = gsub('_', '-', Strain)) %>%
  mutate(Strain = case_when(
    Strain == 'LE17-10' ~ 'LE17-10A',
    Strain == 'LE17-20' ~ 'LE17-20C',
    Strain == 'LE19-282' ~ 'LE19-281',
    Strain == 'LE18-13' ~ 'LE19-13',
    TRUE ~ Strain
  ))  %>%
  # Add back the pseudofeces information
  mutate(pseudofeces = case_when(
      Strain %in% c('LE19-197', 'LE19-196','LE19-59') ~ 'yes',
      TRUE ~ 'no'
    )) %>%
  # These strains were flagged as mixed culture by the Dick lab (source), so their trait might change over time.
  mutate(contamination = case_when(
    Strain %in% c('LE19-12', 'LE19-59') ~ 'yes',
    TRUE ~ 'no'
  )) %>%
  mutate(
    # The amount of Microcystis ingested relative to the preferred food in the mixed treatment
    Unpalatability = as.numeric(blugrn_mix_avg_FI / grn_mix_avg_FI),
    # Total amount phytoplankton ingested in the mixed treatment
    Inhibition = as.numeric(total_chla_mix_avg_FI)) %>%
  select(Strain, pseudofeces, contamination, Unpalatability, Inhibition, blugrn_mix_avg_FI, grn_mix_avg_FI)

# Genome metadata
clade_info <- read.table('inputs/genome_metadata.txt', header = T, sep='\t') %>%
  mutate(Strain = case_when(
    Strain == 'LE17-10' ~ 'LE17-10A',
    Strain == 'LE17-20' ~ 'LE17-20C',
    Strain == 'LE19-282' ~ 'LE19-281',
    Strain == 'LE18-13' ~ 'LE19-13',
    TRUE ~ Strain
  )) %>%
  select(Strain, Clade) 

defense_clade_info <- defense_trait %>%
  left_join(clade_info, by = 'Strain') %>%
  select(Strain, Clade, Unpalatability, Inhibition, pseudofeces, contamination)
```

```{r}
ggplot(defense_clade_info) +
  geom_point(aes(x = Unpalatability, y = Inhibition, fill = Clade, shape = pseudofeces),size = 8, color = "grey30") +
  #The standard FI rate for green algae is 34, which is the average FI of the green alga in the control treatment acclimated to chlamy-only culture in 2025.
  geom_hline(yintercept = 34, color = 'grey30',linetype = 'dashed') +
  annotate("text", x = 0.73, y = 36, hjust = 0,
           label = "Grazing rate \non the preferred food", 
           size = 5, color = "grey10", angle = 0) +
  geom_vline(xintercept = 1, color = 'grey30',linetype = 'dotted') +
  annotate("text", x = 0.95, y = -1.2 , hjust = 0, 
           label = expression("Equal preference \nfor both food sources"), 
           size = 5, color = "grey10", angle = 90) +
  ggrepel::geom_text_repel(aes(label = Strain,x = Unpalatability, y = Inhibition), max.overlaps = 20,size = 2, alpha = 0.8,fontface = 'bold', show.legend = F) +
  scale_fill_manual(values = c('sienna2', 'lightblue2','lightgoldenrodyellow','grey60',
                                 'lightpink4','plum3','palegreen3','lightblue4'),name = expression(bolditalic("Microcystis")~bold("clade"))) +
  scale_shape_manual(values = c(21, 23), name = expression(bold("Expelled food (pseudofeces)"))) +
  #scale_shape_manual(values = c(21, 23), name = 'Non-dissolvable Pseudofeces') +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  labs(title = 'Microcystis Grazing Defense Traits',
       x = 'Mussel selectivity for Microcystis \n (Microcystis : Preferred food)',
       y = 'Total amount of phytoplankton ingested \n (ml/mg AFDW/hr)') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(size=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=16,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=16,color='black',face = "bold"),
        axis.title = element_text(size=10,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=14,color='black',hjust=0.5, face = "bold"),
        legend.text = element_text(size=14,color='black',face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank())
```

# Quantify the effect size of the DAPC-identified BGCs

We wanted to quantify how much the DAPC-identified BGCs contributed to defense relative to each other with a multiple linear regression analysis. Since each *Microcystis* strain has their own unique combination of BGCs, we adopted a targeted approach by using the BGC sequences from one strain as the reference.

## Load reference BGC sequences

The genetic distance to the reference sequence of each BGC is an independent variable whereas the phenotypic value is the dependent variable. To control for the effect of shared evolutionary history, we included a variable of genetic distance based on Microcystis single-copy core genes. The single-copy core genes are shared by at least 98% of the public *Microcystis* pangenome (n = 183; Kiledal et al., 2024).

Loading reference BGC sequences

```{r}
## Each of the vector contains BGC sequences from one strain as the reference. 

### Inhibitory strains
target_clusters_114 <- c("MicroviridinB", "Anabaenopeptin_ctg105","1_heptadecene_ctg44", "Cyanobactin_ctg13", "Cyanobactin_ctg39", "Lanthipeptide_ctg3", "Lanthipeptide_ctg83", "LAP_ctg5","Novel_T3PKS_ctg30", "coregenes")
target_clusters_12 <- c("1_heptadecene_ctg44","MicroviridinB","Anabaenopeptin_ctg105","Novel_T3PKS_ctg30", "NRPS_like_ctg36","Cyanobactin_ctg13", "Cyanobactin_ctg39", "Cyanopeptolin", "Lanthipeptide_ctg3", "Lanthipeptide_ctg83", "LAP_ctg5","RiPP_like_ctg28","coregenes")
target_clusters_84 <- c("1_heptadecene_ctg44","Anabaenopeptin_ctg105","Cyanobactin_ctg13", "Cyanobactin_ctg39","Lanthipeptide_ctg3","Lanthipeptide_ctg83","LAP_ctg5","Novel_T3PKS_ctg30","NRPS_like_ctg36","NRPS_T1PKS_ctg20","Microcystin_LR","Micropeptin_K139","coregenes")

### Unpalatable strains
target_clusters_195 <- c("1_heptadecene_ctg44","MicroviridinB","Anabaenopeptin_ctg105", "Cyanobactin_ctg13","Cyanobactin_ctg39","Lanthipeptide_ctg3","Lanthipeptide_ctg83", "Lanthipeptide_ctg83","LAP_ctg5","Novel_T3PKS_ctg30","NRPS_like_ctg36","coregenes")

all_clusters <- unique(c(target_clusters_114, target_clusters_12, target_clusters_84, target_clusters_195))
## Assign one reference strain
target_clusters <- target_clusters_195
# Get the blast results files
ref_genotype <- 'LE19-195'
```

Get genetic distance to the reference BGC sequences. This is done by blasting the BGC sequences of each strain against the reference BGC sequences. The results are summarized in the `targets_summary.txt` file, which contains the sequence identity of each BGC sequence to the reference sequence. The sequence identity is used as a measure of genetic distance.

```{r, message=FALSE}
bgc_blast_results <- file.path('inputs/curated_bgc_genotype',ref_genotype)
## list all directories under the bgc_blast_results
bgc_dirs <- list.dirs(bgc_blast_results, recursive = FALSE) %>%
  .[!grepl('miscallaneous',.)] # exclude the fragmented and not well-defined bgcs
## find all blast result files under each directory
bgc_seqidt_files <- unlist(lapply(bgc_dirs, function(dir) {
  list.files(file.path(dir,'blast_results/'), pattern = 'targets_summary.txt', full.names = TRUE)
}))

# Get core gene blast results
core_blast_results <- file.path('inputs/reference_core_genes',ref_genotype,'blast_results/')
core_seqidt_files <- list.files(core_blast_results, pattern = 'targets_summary.txt', full.names = TRUE)

# Combine the BGC and core gene blast results
seqidt_files <- c(bgc_seqidt_files, core_seqidt_files)
gene_names <- gsub(paste0(ref_genotype,'_|_targets_summary.txt'), '', basename(seqidt_files))
```

Compute distance matrix for each BGC and core gene. 

```{r}
# Load sequence identity data for each BGC and core gene
seq_sim_list = list()

for (clusters in seq_along(seqidt_files)){
  cluster_name <- basename(seqidt_files[clusters]) %>%
    str_extract('(?<=_).*(?=_targets_summary.txt)')
  seq_idt_table <- read_seq_identity(seqidt_files[clusters]) %>%
    select(Strain, ANI, Percent_presence) %>%
    mutate(cluster = cluster_name)
  seq_sim_list[[clusters]] <- seq_idt_table
}

seq_sim_table <- bind_rows(seq_sim_list) %>%
  pivot_longer(cols = c(ANI, Percent_presence), names_to = 'Seq_Metric', values_to = 'Value') %>%
  drop_na()

# tidy up the cluster names
seq_sim_table <- seq_sim_table %>%
  mutate(cluster = case_when(
    cluster == 'coregenes' ~ 'Core Genes',
    cluster == '1_heptadecene_ctg44' ~ '1-Heptadecene',
    cluster == 'Anabaenopeptin_ctg105' ~ 'Anabaenopeptin',
    cluster == 'Cyanobactin_ctg13' ~ 'Cyanobactin (ctg13)',
    cluster == 'Cyanobactin_ctg39' ~ 'Cyanobactin (ctg39)',
    cluster == 'Lanthipeptide_ctg3' ~ 'Lanthipeptide (ctg3)',
    cluster == 'Lanthipeptide_ctg83' ~ 'Lanthipeptide (ctg83)',
    cluster == 'LAP_ctg5' ~ 'LAP (ctg5)',
    cluster == 'Novel_T3PKS_ctg30' ~ 'Novel-T3PKS (ctg30)',
    cluster == 'NRPS_like_ctg36' ~ 'NRPS-like (ctg36)',
    cluster == 'NRPS_T1PKS_ctg20' ~ 'NRPS-T1PKS (ctg20)',
    cluster == 'Microcystin_LR' ~ 'Microcystin',
    cluster == 'Micropeptin_K139' ~ 'Micropeptin',
    cluster == 'MicroviridinB' ~ 'Microviridin',
    cluster == 'RiPP_like_ctg28' ~ 'RiPP-like (ctg28)',
    TRUE ~ cluster
  ))
clean_cluster_names <- as.data.frame(target_clusters) %>%
  mutate(target_clusters = case_when(
    target_clusters == 'coregenes' ~ 'Core Genes',
    target_clusters == '1_heptadecene_ctg44' ~ '1-Heptadecene',
    target_clusters == 'Anabaenopeptin_ctg105' ~ 'Anabaenopeptin',
    target_clusters == 'Cyanobactin_ctg13' ~ 'Cyanobactin (ctg13)',
    target_clusters == 'Cyanobactin_ctg39' ~ 'Cyanobactin (ctg39)',
    target_clusters == 'Lanthipeptide_ctg3' ~ 'Lanthipeptide (ctg3)',
    target_clusters == 'Lanthipeptide_ctg83' ~ 'Lanthipeptide (ctg83)',
    target_clusters == 'LAP_ctg5' ~ 'LAP (ctg5)',
    target_clusters == 'Novel_T3PKS_ctg30' ~ 'Novel-T3PKS (ctg30)',
    target_clusters == 'NRPS_like_ctg36' ~ 'NRPS-like (ctg36)',
    target_clusters == 'NRPS_T1PKS_ctg20' ~ 'NRPS-T1PKS (ctg20)',
    target_clusters == 'Microcystin_LR' ~ 'Microcystin',
    target_clusters == 'Micropeptin_K139' ~ 'Micropeptin',
    target_clusters == 'MicroviridinB' ~ 'Microviridin',
    target_clusters == 'RiPP_like_ctg28' ~ 'RiPP-like (ctg28)',
    TRUE ~ target_clusters
  )) %>%
  pull()
```

```{r}
## Genetic distance matrix 1: Average nucleotide identity (ANI) among shared genes to the reference BGC sequences
ANI_dis_table <- seq_sim_table %>%
  filter(Seq_Metric == 'ANI') %>%
  select(Strain, cluster, Value) %>%
  pivot_wider(names_from = cluster, values_from = Value) %>%
  left_join(defense_clade_info, by = 'Strain')%>%
  drop_na()

## Genetic distance matrix 2: Percentage of the reference BGC genes present in each strain
pres_dis_table <- seq_sim_table %>%
  filter(Seq_Metric == 'Percent_presence') %>%
  select(Strain, cluster, Value) %>%
  pivot_wider(names_from = cluster, values_from = Value) %>%
  left_join(defense_clade_info, by = 'Strain') %>%
  drop_na()
```

## Multiple Linear Regression Analysis

### Genetic predictor: ANI

Data transformation to meet assumptions

```{r}
# Prepare the data for multivariate multiple regression
mvlm_ani_table <- ANI_dis_table %>% 
  mutate(log_inhib = log(Inhibition)) %>%
  # scale both outcomes so that the prediction results can be comparable
  mutate(std_unpal = scale(Unpalatability), std_log_inhib = scale(log_inhib)) %>%
  select(all_of(clean_cluster_names),std_unpal,std_log_inhib)
sample_size <- nrow(mvlm_ani_table)
```

#### Model comparisons and MLRs

```{r}
## Perform step-wise selection
# BIC-based method given small sample size
ani_model_inhib <- lm(std_log_inhib ~ ., data = mvlm_ani_table %>% select(-std_unpal) )
ani_model_unpal <- lm(std_unpal ~ ., data = mvlm_ani_table %>% select(-std_log_inhib) )
## stepwise regression for model selection
step_model_ani_inhib <- step(ani_model_inhib, direction = "both", scope = list(lower = ~ `Core Genes`), k = log(sample_size))
step_model_ani_unpal <- step(ani_model_unpal, direction = "both", scope = list(lower = ~ `Core Genes`), k = log(sample_size))
summary(step_model_ani_inhib); hist(step_model_ani_inhib$residuals); shapiro.test(step_model_ani_inhib$residuals)
summary(step_model_ani_unpal); hist(step_model_ani_unpal$residuals); shapiro.test(step_model_ani_unpal$residuals)
```

#### Assumption check

```{r}

plot(step_model_ani_inhib, which = 1, main = "Trait: Inhibition") # Residuals vs Fitted
plot(step_model_ani_inhib, which = 2, main = "Trait: Inhibition") # Normal Q-Q
plot(step_model_ani_unpal, which = 1, main = "Trait: Unpalatability") # Residuals vs Fitted)
plot(step_model_ani_unpal, which = 2, main = "Trait: Unpalatability") # Normal Q-Q

```

#### Record model fit

```{r}
ani_model_annotations <- bind_rows(
  broom::glance(step_model_ani_inhib) %>% mutate(Outcome = "Log Inhibition"),
  broom::glance(step_model_ani_unpal) %>% mutate(Outcome = "Unpalatability")
) %>%
  mutate(Gen_metric = "Shared gene ANI") %>%
  mutate(
    annotation = paste0(
      "Adj. R² = ", round(adj.r.squared, 3), "\n",
      "F(", df, ",", df.residual, ") = ", round(statistic, 2), "\n",
      case_when(
        p.value < 0.001 ~ "p < 0.001 ***",
        p.value < 0.01 ~ paste0("p = ", round(p.value, 3), " **"),
        p.value < 0.05 ~ paste0("p = ", round(p.value, 3), " *"),
        p.value < 0.1 ~ paste0("p = ", round(p.value, 3), " ."),
        TRUE ~ paste0("   p = ", round(p.value, 3))
      )
    )
  )
```

#### Reporting coefficients and forest plot

```{r}
# Get bootstrap confidence intervals for the stepwise model
tidy_inhib <- tidy(step_model_ani_inhib, conf.int = TRUE) %>%
  mutate(Outcome = "Log Inhibition")
tidy_unpal <- tidy(step_model_ani_unpal, conf.int = TRUE) %>%
  mutate(Outcome = "Unpalatability")

combined_ani_boot <- bind_rows(tidy_inhib,tidy_unpal) %>%
  filter(term != "(Intercept)") %>%
  mutate(Significant = case_when(
    conf.low > 0 ~ "Reduced defense",
    conf.high < 0 ~ "Stronger defense",
    TRUE ~ "No significant effect"
  ),
  term = gsub("`", "", term),
  term = gsub("std_", "", term),
  effect_cat = case_when(
    abs(estimate) > 20 ~ "Strong effect",
    abs(estimate) <= 20 ~ "Small effect",
  ),
  Ref_genotype = ref_genotype,
  Gen_metric = "Shared gene ANI"
)
write.csv(combined_ani_boot, paste0("output_tables/coefficients/",ref_genotype,"_ani_model_coefficients.csv"), row.names = FALSE)
```

### Genetic predictor: Gene content conservation

Data transformation to meet assumptions

```{r}
# Prepare the data for multivariate multiple regression
mvlm_pres_table <- pres_dis_table %>% 
  mutate(log_inhib = log(Inhibition)) %>%
  # scale both outcomes so that the prediction results can be comparable
  mutate(std_unpal = scale(Unpalatability), std_log_inhib = scale(log_inhib)) %>%
  select(all_of(clean_cluster_names),std_unpal,std_log_inhib)
sample_size <- nrow(mvlm_pres_table)
```

#### Model comparisons and MLRs

```{r}
## Perform step-wise selection
# BIC-based method given small sample size
pres_model_inhib <- lm(std_log_inhib ~ ., data = mvlm_pres_table %>% select(-std_unpal) )
pres_model_unpal <- lm(std_unpal ~ ., data = mvlm_pres_table %>% select(-std_log_inhib) )
## stepwise regression for model selection
step_model_pres_inhib <- step(pres_model_inhib, direction = "both", scope = list(lower = ~ `Core Genes`), k = log(sample_size))
step_model_pres_unpal <- step(pres_model_unpal, direction = "both", scope = list(lower = ~ `Core Genes`), k = log(sample_size))
summary(step_model_pres_inhib); hist(step_model_pres_inhib$residuals); shapiro.test(step_model_pres_inhib$residuals)
summary(step_model_pres_unpal); hist(step_model_pres_unpal$residuals); shapiro.test(step_model_pres_unpal$residuals)
```

#### Assumption check

```{r}

plot(step_model_pres_inhib, which = 1, main = "Trait: Inhibition") # Residuals vs Fitted
plot(step_model_pres_inhib, which = 2, main = "Trait: Inhibition") # Normal Q-Q
plot(step_model_pres_unpal, which = 1, main = "Trait: Unpalatability") # Residuals vs Fitted)
plot(step_model_pres_unpal, which = 2, main = "Trait: Unpalatability") # Normal Q-Q

```

#### Record model fit

```{r}
pres_model_annotations <- bind_rows(
  broom::glance(step_model_pres_inhib) %>% mutate(Outcome = "Log Inhibition"),
  broom::glance(step_model_pres_unpal) %>% mutate(Outcome = "Unpalatability")
) %>%
  mutate(Gen_metric = "Gene content conservation") %>%
  mutate(
    annotation = paste0(
      "Adj. R² = ", round(adj.r.squared, 3), "\n",
      "F(", df, ",", df.residual, ") = ", round(statistic, 2), "\n",
      case_when(
        p.value < 0.001 ~ "p < 0.001 ***",
        p.value < 0.01 ~ paste0("p = ", round(p.value, 3), " **"),
        p.value < 0.05 ~ paste0("p = ", round(p.value, 3), " *"),
        p.value < 0.1 ~ paste0("p = ", round(p.value, 3), " ."),
        TRUE ~ paste0("  p = ", round(p.value, 3))
      )
    )
  )
```

#### Reporting coefficients and forest plot

```{r}
# Get bootstrap confidence intervals for the stepwise model
tidy_inhib <- tidy(step_model_pres_inhib, conf.int = TRUE) %>%
  mutate(Outcome = "Log Inhibition")
tidy_unpal <- tidy(step_model_pres_unpal, conf.int = TRUE) %>%
  mutate(Outcome = "Unpalatability")

combined_pres_boot <- bind_rows(tidy_inhib,tidy_unpal) %>%
  filter(term != "(Intercept)") %>%
  mutate(Significant = case_when(
    conf.low > 0 ~ "Reduced defense",
    conf.high < 0 ~ "Stronger defense",
    TRUE ~ "No significant effect"
  ),
  term = gsub("`", "", term),
  term = gsub("std_", "", term),
  effect_cat = case_when(
    abs(estimate) > 20 ~ "Strong effect",
    abs(estimate) <= 20 ~ "Small effect",
  ),
  Ref_genotype = ref_genotype,
  Gen_metric = "Gene content conservation"
)
write.csv(combined_pres_boot, paste0("output_tables/coefficients/",ref_genotype,"_pres_model_coefficients.csv"), row.names = FALSE)
```

# Plot the effect sizes of the BGCs identified by DAPC

```{r}
forest_table <- bind_rows(combined_ani_boot, combined_pres_boot) %>%
  mutate(Gen_metric = factor(Gen_metric, levels = c("Shared gene ANI", "Gene content conservation")))

annotation_table <- bind_rows(ani_model_annotations, pres_model_annotations)
```

## Inhibition

```{r}
## forest plot of the model coefficients for both outcomes
ggplot(forest_table %>%
         filter(Outcome =="Log Inhibition") ,
       aes(x = estimate, y = reorder(term, estimate), color = Significant)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    position = position_dodge(width = 0.6),
    height = 0.3
  ) +
  scale_color_manual(values = c("Reduced defense" = "indianred", 
                                "Stronger defense" = "steelblue", 
                                "No significant effect" = "gray")) +
  geom_text(
    data = annotation_table %>%
      filter(Outcome == "Log Inhibition"),
    aes(x = Inf, y = Inf, label = annotation),
    inherit.aes = FALSE,
    vjust = 8,
    hjust = 1.2,
    size = 2.5,
    color = "black"
  ) +
  facet_wrap(~ Gen_metric, scales = "free_x") +
  labs(
    title    = "Response variable: Inhibition",
    x        = "Effect size (95% CI)",
    y        = "Biosynthetic Gene Cluster",
    color    = "Direction of Effect"
  ) +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.5,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=12,color='black',face = "bold"),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
        strip.text = element_text(size = 12, color = 'black',face = "bold"),
        strip.background = element_rect(fill = "grey100", 
                                        color = NA, linewidth = 1),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=12,color='black',hjust=0, face = "bold"),
        legend.text = element_text(size=10,color='black',hjust=0),
        legend.position = 'right',
        panel.grid.major = element_blank())
ggsave(paste0("output_tables/forest_plots/",ref_genotype,"_inhibition_model_coefficients_forest_plot.png"), width = 10, height = 4)
```

## Unpalatability

```{r}
## forest plot of the model coefficients for both outcomes
ggplot(forest_table %>%
         filter(Outcome =="Unpalatability") ,
       aes(x = estimate, y = reorder(term, estimate), color = Significant)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    position = position_dodge(width = 0.6),
    height = 0.3
  ) +
  scale_color_manual(values = c("Reduced defense" = "indianred", 
                                "Stronger defense" = "steelblue", 
                                "No significant effect" = "gray")) +
  geom_text(
    data = annotation_table %>%
      filter(Outcome == "Unpalatability"),
    aes(x = Inf, y = Inf, label = annotation),
    inherit.aes = FALSE,
    vjust = 8,
    hjust = 1.2,
    size = 2.5,
    color = "black"
  ) +
  facet_wrap(~ Gen_metric, scales = "free_x") +
  labs(
    title    = "Response variable: Unpalatability",
    x        = "Effect size (95% CI)",
    y        = "Biosynthetic Gene Cluster",
    color    = "Direction of Effect"
  ) +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.5,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=12,color='black',face = "bold"),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
        strip.text = element_text(size = 12, color = 'black',face = "bold"),
        strip.background = element_rect(fill = "grey100", 
                                        color = NA, linewidth = 1),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=12,color='black',hjust=0, face = "bold"),
        legend.text = element_text(size=10,color='black',hjust=0),
        legend.position = 'right',
        panel.grid.major = element_blank())
ggsave(paste0("output_tables/forest_plots/",ref_genotype,"_unpalatability_model_coefficients_forest_plot.png"), width = 10, height = 4)
```