---
title: "fig2_bgc_trait_assoc"
author: "Yu-Cheng Lin"
date: "2026-05-21"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)
library(adegenet)
source('scripts/parse_functions.R')
```

# Defense trait

Load defense trait data

```{r}
# Trait data summary from dreissinid mussel feeding experiments
raw_defense_trait <- read.table('inputs/LE_QMF_23_IDM_trait_summary.out.txt', header = T, sep = '\t')
defense_trait <- raw_defense_trait%>%
  mutate(Strain = str_extract(Culture, '^[^.]+')) %>%
  mutate(Strain = gsub('_', '-', Strain)) %>%
  mutate(Strain = case_when(
    Strain == 'LE17-10' ~ 'LE17-10A',
    Strain == 'LE17-20' ~ 'LE17-20C',
    Strain == 'LE19-282' ~ 'LE19-281',
    Strain == 'LE18-13' ~ 'LE19-13',
    TRUE ~ Strain
  ))  %>%
  # Add back the pseudofeces information
  mutate(pseudofeces = case_when(
      Strain %in% c('LE19-197', 'LE19-196','LE19-59') ~ 'yes',
      TRUE ~ 'no'
    )) %>%
  # These strains were flagged as mixed culture by the Dick lab (source), so their trait might change over time.
  mutate(contamination = case_when(
    Strain %in% c('LE19-12', 'LE19-59') ~ 'yes',
    TRUE ~ 'no'
  )) %>%
  mutate(
    # The amount of Microcystis ingested relative to the preferred food in the mixed treatment
    Unpalatability = as.numeric(blugrn_mix_avg_FI / grn_mix_avg_FI),
    # Total amount phytoplankton ingested in the mixed treatment
    Inhibition = as.numeric(total_chla_mix_avg_FI)) %>%
  select(Strain, pseudofeces, contamination, Unpalatability, Inhibition, blugrn_mix_avg_FI, grn_mix_avg_FI)

# Genome metadata
clade_info <- read.table('inputs/genome_metadata.txt', header = T, sep='\t') %>%
  mutate(Strain = case_when(
    Strain == 'LE17-10' ~ 'LE17-10A',
    Strain == 'LE17-20' ~ 'LE17-20C',
    Strain == 'LE19-282' ~ 'LE19-281',
    Strain == 'LE18-13' ~ 'LE19-13',
    TRUE ~ Strain
  )) %>%
  select(Strain, Clade) 

defense_clade_info <- defense_trait %>%
  left_join(clade_info, by = 'Strain') %>%
  select(Strain, Clade, Unpalatability, Inhibition, pseudofeces, contamination)
```

```{r}
ggplot(defense_clade_info) +
  geom_point(aes(x = Unpalatability, y = Inhibition, fill = Clade, shape = pseudofeces),size = 8, color = "grey30") +
  #The standard FI rate for green algae is 34, which is the average FI of the green alga in the control treatment acclimated to chlamy-only culture in 2025.
  geom_hline(yintercept = 34, color = 'grey30',linetype = 'dashed') +
  annotate("text", x = 0.73, y = 36, hjust = 0,
           label = "Grazing rate \non the preferred food", 
           size = 5, color = "grey10", angle = 0) +
  geom_vline(xintercept = 1, color = 'grey30',linetype = 'dotted') +
  annotate("text", x = 0.95, y = -1.2 , hjust = 0, 
           label = expression("Equal preference \nfor both food sources"), 
           size = 5, color = "grey10", angle = 90) +
  #ggrepel::geom_text_repel(aes(label = Strain,x = Unpalatability, y = Inhibition), max.overlaps = 20,size = 2, alpha = 0.8,fontface = 'bold', show.legend = F) +
  scale_fill_manual(values = c('sienna2', 'lightblue2','lightgoldenrodyellow','grey60',
                                 'lightpink4','plum3','palegreen3','lightblue4'),name = expression(bolditalic("Microcystis")~bold("clade"))) +
  scale_shape_manual(values = c(21, 23), name = expression(bold("Expelled food (pseudofeces)"))) +
  #scale_shape_manual(values = c(21, 23), name = 'Non-dissolvable Pseudofeces') +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  labs(title = 'Two Dimensions of Microcystis Grazing Defense Traits',
       x = 'Mussel selectivity for Microcystis \n (Microcystis : Preferred food)',
       y = 'Total amount of phytoplankton ingested \n (ml/mg AFDW/hr)') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(size=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=16,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=16,color='black',face = "bold"),
        axis.title = element_text(size=10,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=14,color='black',hjust=0.5, face = "bold"),
        legend.text = element_text(size=14,color='black',face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank())
```

# Variations in Biosynthetic gene clusters

## Process annotation

The nucleotide sequences of the 21 Microcystis Lake Erie (LE) strains were annotated by a curated BGC database (Hart et al., 2026, ISME) using blastn (see Material and Methods in the Supplementary Material). The BGC database file is located in `inputs/annotated_bgc_genes/`` under the name `BGCdb_seed5_key.csv`. The annotated results were located in `inputs/annotated_bgc_genes/` under the name of `_lauren_blastn_table.csv`. Here, we combine the annotation results into a single table.

```{r load_data}
# Set directory
annot_result_path <- 'inputs/annotated_bgc_genes/'

# Load BGC keys for the curated database
bgc_key <- read.csv(paste0(annot_result_path,'BGCdb_seed5_key.csv'), header = TRUE) %>%
  select(Header, Main_Product, Biosynthetic_Class, Organism, Gene, New_BGC_name,
         gene_type, pos_in_bgc, core_genes, genes_in_bgc)

# Find the annotation results in the format of LEXXXX_lauren_blastn_table.csv
pattern <- '_lauren_blastn_table.csv$'
anno_dirs <- list.files(annot_result_path, pattern = pattern, full.names = TRUE)
parse_single_anno_result(anno_dirs[1],bgc_key)

# Parse multiple annotation results and combine with annotation key
bgc_annot_combined <- parse_multiple_anno_results(anno_dirs, bgc_key)

# Output table
write.csv(bgc_annot_combined, 'output_tables/annotated_BGCs_lauren_microcystis_LE_strains.csv', row.names = FALSE)

```

## Calculate shared gene ANI and gene content conservation

```{r}

#bgc_annot_combined <- read.csv('output_tables/annotated_BGCs_lauren_microcystis_LE_strains.csv', header = T)

# Calculate genetic variation metrics
bgc_ani_pres <- bgc_annot_combined %>%
  # Cleaner names
  mutate(cluster = case_when(
    New_BGC_name == 'Cyanobactin-Microcystis aeruginosa WLECC' & grepl('ctg13', Gene) ~'Cyanobactin ctg13',
    New_BGC_name == 'Cyanobactin-Microcystis aeruginosa WLECC' & grepl('ctg39', Gene) ~'Cyanobactin ctg39',
    New_BGC_name == 'Cyanobactin-Microcystis aeruginosa WLECC' & grepl('ctg115', Gene) ~'Cyanobactin ctg115',
    grepl('Microcystin LR-Microcystis', New_BGC_name) ~ 'Microcystin-Microcystis',
    grepl('Microviridin B-Microcystis', New_BGC_name) ~ 'Microviridin-Microcystis',
    TRUE ~ New_BGC_name
  )) %>%
  mutate(cluster = gsub(cluster, 
                        pattern = '-Microcystis aeruginosa WLECC', 
                        replacement = '')) %>%
  mutate(cluster = gsub(cluster, 
                        pattern = '-Microcystis', 
                        replacement = '')) %>%
  mutate(cluster = gsub(cluster, 
                        pattern = 'lanthipeptide', 
                        replacement = 'Lanthipeptide')) %>%
  # Calculate ANI and presence absence
  select(Strain,cluster, Gene, gene_type, Identity_Score) %>%
  filter(cluster != 'NA') %>%
  mutate(ANI = case_when(
    is.na(Identity_Score) ~ NA,
    TRUE ~ Identity_Score
  )) %>%
  mutate(Presence = case_when(
    is.na(Identity_Score) ~ 0,
    TRUE ~ 1
  )) %>%
  mutate(core_presence = case_when(
    gene_type == 'core' & Presence == 1 ~ 1,
    gene_type == 'core' & Presence == 0 ~ 0,
    TRUE ~ NA_real_
  )) %>%
  group_by(Strain, cluster) %>%
  summarise(ANI = mean(ANI, na.rm = T),
            Percent_presence = mean(Presence, na.rm = T)*100,
            Core_gene_presence = mean(core_presence, na.rm = T))

# Merge genetic variation and trait data
bgc_trait <- bgc_ani_pres %>%
  inner_join(defense_clade_info, by = c('Strain' = 'Strain'))
```

## Quality control

```{r}
# Total number of BGCs in the database
length(unique(bgc_trait$cluster))
```

```{r}
bgc_trait_95 <- bgc_trait %>%
  filter(!is.na(cluster)) %>%
mutate(distant_cluster = case_when(
    ANI > 95 ~ cluster,
    TRUE ~ NA
  ))
# Relationship between the nucleotide divergence and presence absence
ani_pres_plot <- ggplot(bgc_trait_95, aes(x = ANI, y = Percent_presence, color = distant_cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = 'BGC variation among Microcystis WLECC strains',
       x = 'Shared gene average \nnucleotide identity (ANI%)',
       y = 'Gene content conservation \n(percent presence %)',
       color = 'BGC Gene Cluster') +
  geom_vline(xintercept = 95, linetype = 'dashed', color = 'grey50') +
  geom_hline(yintercept = 17, linetype = 'dashed', color = 'grey50') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold"),
        axis.title = element_text(size=15,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'NA',
        panel.grid.major = element_blank())
ggsave('figures/supplFig1A_rawbgc_variation.png', ani_pres_plot, width = 6, height = 6, dpi = 300)
ani_pres_plot
```

```{r}
# boxplot show variation among strains
boxplot_bgc_gene_ani <- ggplot(bgc_trait_95, aes(x = ANI, y = reorder(cluster, Percent_presence, decreasing = T))) +
  geom_boxplot(fill = 'grey') +
  labs(title = 'Shared Gene ANI% variation in Microcystis WLECC',
       x = 'Average nucleotide identity (ANI%)',
       y = 'BGC Cluster Type') +
  geom_vline(xintercept = 95, linetype = 'dashed', color = 'grey50') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=8,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=15,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank()) 
boxplot_bgc_gene_pres <- ggplot(bgc_trait_95, aes(x = Percent_presence, y = reorder(cluster, Percent_presence, decreasing = T))) +
  geom_boxplot(fill = 'grey') +
  labs(title = 'Gene content conservation variation in Microcystis WLECC',
       x = 'Gene content conservation (%)',
       y = 'BGC Cluster Type') +
  theme_pubclean()+
  theme(plot.title= element_text(linewidth=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=12,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold"),
        axis.title = element_text(size=15,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank())
  
ggsave('figures/supplFig1B_ani_boxplot.png', boxplot_bgc_gene_ani, width = 6, height = 6, dpi = 300)
ggsave('figures/supplFig1B_gene_content_boxplot.png', boxplot_bgc_gene_pres, width = 6, height = 6, dpi = 300)
boxplot_bgc_gene_ani; boxplot_bgc_gene_pres
```

Some clusters are rare in the strain collection.

```{r}
# Cleaning of rare clusters that are not present in more than 70% of the strains, which might be less representative.
rare_cluster <- bgc_trait %>%
  group_by(cluster) %>%
  summarise(
    median_presence = median(Percent_presence, na.rm = T),
    zero_count = sum(Percent_presence == 0),
    total_count = n()) %>%
  mutate(zero_percentage = zero_count / total_count) %>%
  drop_na() %>%
  filter(zero_percentage > .7)
## Total number of rare BGCs
nrow(rare_cluster)
```

### Cutoff determination for removing rare clusters

We will plot proportion of presence in strains and the median of percent gene presence for each cluster to determine the cutoff for removing rare and incomplete clusters.

```{r}
# summarize the zero percentage and median of percent gene presence for each cluster
cluster_quality <- bgc_trait %>%
  group_by(cluster) %>%
  summarise(
    median_presence = median(Percent_presence, na.rm = T),
    zero_count = sum(Percent_presence == 0),
    total_count = n()) %>%
  mutate(presence_percentage = (1 - (zero_count / total_count)) *100) 

# plot a scatter plot for both metrics
ggplot(cluster_quality, aes(x = median_presence, y = presence_percentage)) +
  geom_point(color = 'indianred3', size = 4) +
  ggrepel::geom_text_repel(aes(label = cluster), max.overlaps = 10,
                     size = 2, alpha = 0.8,fontface = 'bold', show.legend = F) +
  geom_hline(yintercept = 30, linetype = 'dashed', color = 'grey50') +
  annotate("text", x = 25, y = 35, hjust = 0,
           label = "30% presence in strains", 
           size = 5, color = "grey30", angle = 0) +
  geom_vline(xintercept = 17, linetype = 'dotted', color = 'grey50') +
  annotate("text", x = 18, y = 10, hjust = 0,
           label = "17% median gene presence", 
           size = 5, color = "grey30", angle = 0) +
  scale_x_continuous(limits = c(0, 105)) +
  scale_y_continuous(limits = c(0, 105)) +
  labs(title = 'BGC quality metrics',
       x = 'Median percent gene presence',
       y = 'Proportion of presence in strains') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold"),
        axis.title = element_text(size=15,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'NA',
        panel.grid.major = element_blank())

ggsave('figures/supplFig1C_bgc_quality_metrics.png', width = 8, height = 6, dpi = 300)
```


We will plot a bar graph of the number of BGCs kept at each cutoff.

```{r}
cutoff_values <- seq(0, 1, by = 0.05)

cutoff_results <- data.frame(cutoff = numeric(), num_clusters = numeric())

for (cutoff in cutoff_values) {
  num_clusters <- bgc_trait %>%
    group_by(cluster) %>%
    summarise(zero_count = sum(Percent_presence == 0),
              total_count = n()) %>%
    mutate(zero_percentage = zero_count / total_count) %>%
    filter(zero_percentage <= cutoff) %>%
    nrow()
  
  cutoff_results <- rbind(cutoff_results, data.frame(cutoff = cutoff, num_clusters = num_clusters))
}

# Plot the results

ggplot(cutoff_results, aes(x = cutoff, y = num_clusters)) +
  geom_line(color = 'cyan4') +
  geom_point(color = 'indianred3') +
  labs(title = 'Number of BGCs kept at different cutoffs',
       x = 'Cutoff for zero percentage',
       y = 'Number of BGCs retained') +
  theme_pubclean()

```


```{r}
bgc_trait_clean <- bgc_trait %>%
  filter(!cluster %in% rare_cluster$cluster)
```

The low gene content conservation and ANI values for some clusters suggested that they might be false positive hits. The clusters may have extensive gene loss in these tested strains or the hits are resulted from off-target annotation due to the poor reference sequence. To avoid the noise from these clusters, we removed the clusters with low percent presence for the downstream analysis.

```{r}
## Identify the clusters that show low percent presence among strains, which might be less representative.
low_present_cluster <- bgc_trait_clean %>%
  group_by(cluster) %>%
  summarise(
    mean_presence = mean(Percent_presence, na.rm = T),
    median_presence = median(Percent_presence, na.rm = T),
    zero_count = sum(Percent_presence == 0),
    total_count = n()) %>%
  mutate(zero_percentage = zero_count / total_count) %>%
  filter(median_presence < 17) %>%
  pull(cluster)

length(low_present_cluster)
```

```{r}
bgc_trait_dbclean <- bgc_trait_clean %>%
  filter(!cluster %in% low_present_cluster)%>%
mutate(distant_cluster = case_when(
    ANI > 95 ~ cluster,
    TRUE ~ NA
  ))
```

```{r}
# list the remaining clusters
bgc_trait_dbclean %>%
  group_by(cluster) %>%
  summarise(median_presence = median(Percent_presence, na.rm = T),
            zero_count = sum(Percent_presence == 0),
            total_count = n()) %>%
  mutate(zero_percentage = zero_count / total_count) %>%
  arrange(desc(median_presence)) %>%
  write.csv('output_tables/filtered_bgc_variation_summary.csv', row.names = FALSE)
```

After filtering

```{r}
# Genetic variation after filtering
ani_pres_clean_plot <- ggplot(bgc_trait_dbclean, aes(x = ANI, y = Percent_presence, color = distant_cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = 'Filtered BGC variation among Microcystis WLECC strains',
       x = 'Shared gene average nucleotide identity (ANI%)',
       y = 'Gene content conservation (percent presence %)',
       color = 'BGC Gene Cluster') +
  geom_vline(xintercept = 95, linetype = 'dashed', color = 'grey50') +
  geom_hline(yintercept = 17, linetype = 'dashed', color = 'grey50') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold"),
        axis.title = element_text(size=15,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'NA',
        panel.grid.major = element_blank())
ggsave('figures/supplFig1A_cleanbgc_variation.png', ani_pres_clean_plot, width = 6, height = 6, dpi = 300)
ani_pres_clean_plot

```

## Prepare data for DAPC

```{r}
## Remove the clusters with low percent presence
bgc_pres_wide <- bgc_trait_dbclean %>%
  filter(!cluster %in% low_present_cluster) %>%
  select(Strain, cluster, Percent_presence) %>%
  pivot_wider(id_cols = Strain, names_from = cluster, values_from = Percent_presence, values_fill = list(Percent_presence = 0)) %>%
  ungroup() 

bgc_ani_wide <- bgc_trait_dbclean %>%
  filter(!cluster %in% low_present_cluster) %>%
  select(Strain, cluster, ANI) %>%
  mutate(ANI = case_when(
    is.na(ANI) ~ 0,
    TRUE ~ ANI
  )) %>%
  pivot_wider(id_cols = Strain, names_from = cluster, values_from = ANI, values_fill = list(ANI = 0)) %>%
  ungroup()
```

# Gene-trait association using Discriminant Analysis of Principal Components (DAPC)

## Phenotypic groupings based on hierarchical clustering

```{r}
inhibitory_strains <- c("LE17-10A","LE19-114","LE19-4")

# Define unpalatability groupings
unpala_clusters <- defense_clade_info %>%
  select(Strain, Unpalatability) %>%
  # Exclude the inhibitory strains to ensure mazimized detection of genetic features contributing to unpalatability
  filter(!Strain %in%inhibitory_strains) %>%
  column_to_rownames(var = 'Strain') %>%
  scale() %>%
  dist() %>%
  hclust(method = 'ward.D2')

#dendrograms

plot(unpala_clusters, main = 'Hierarchical Clustering of Unpalatability', xlab = '', sub = '', cex = 0.6)
rect.hclust(unpala_clusters, k = 4, border = 'indianred')

```

```{r}
# Define inhibition groupings
inhib_clusters <- defense_clade_info %>%
  select(Strain, Inhibition) %>%
  column_to_rownames(var = 'Strain') %>%
  scale() %>%
  dist() %>%
  hclust(method = 'ward.D2')
#dendrograms
plot(inhib_clusters, main = 'Hierarchical Clustering of Inhibition', xlab = '', sub = '', cex = 0.6)
rect.hclust(inhib_clusters, k = 3, border = 'cyan4')
```

Plot hc groups on the trait distribution plot

```{r}
defense_trait_DAPC <- defense_clade_info %>%
  mutate(unpala_hc_groups = as.factor(cutree(unpala_clusters, k = 4)[Strain])) %>%
  mutate(inhib_hc_groups = as.factor(cutree(inhib_clusters, k = 3)[Strain])) %>%
  mutate(unpala_hc_groups = case_when(
    is.na(unpala_hc_groups) ~ 'Not included',
    TRUE ~ as.character(unpala_hc_groups)
  )) %>%
  mutate(arrang_unpala_hc_groups = case_when(
    unpala_hc_groups == '1' ~ '2',
    unpala_hc_groups == '2' ~ '3',
    unpala_hc_groups == '3' ~ '4',
    unpala_hc_groups == '4' ~ '1',
    unpala_hc_groups == 'Not included' ~ 'grey70'
  ))

## plot unpalatability
ggplot(defense_trait_DAPC, 
       aes(x = Unpalatability, y = Inhibition,
           color = arrang_unpala_hc_groups)) +
  geom_point(size = 6) +
  #The standard FI rate for green algae is 34, which is the average FI of the green alga in the control treatment acclimated to chlamy-only culture in 2025.
  geom_hline(yintercept = 34, color = 'grey30',linetype = 'dashed') +
  geom_vline(xintercept = 1, color = 'grey30',linetype = 'dotted') +
  #ggrepel::geom_text_repel(aes(label = Strain), max.overlaps = 20,
  #                           size = 2, alpha = 0.8,fontface = 'bold', show.legend = F) +
  scale_color_manual(
    values = c('lightpink4','rosybrown1','darkseagreen3','aquamarine4',"grey70"),
    name = paste0('Unpalatability Groups',"\n",'(Hierarchical Clustering)')) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  labs(title = 'Defense traits of Microcystis WLECC strains',
       x = 'Mussel selectivity for Microcystis (Unpalatability)',
       y = 'Total amount of phytoplankton ingested by mussel (Inhibition)') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth =0, colour = "black"),
        axis.ticks = element_line(linewidth =1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=16,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=16,color='black',face = "bold"),
        axis.title = element_text(size=10,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank())
ggsave('figures/fig2A_unpala_hc_groups.png', width = 6, height = 4, dpi = 300)

```

```{r}
# plot inhibition
ggplot(defense_trait_DAPC, 
       aes(x = Unpalatability, y = Inhibition,color = inhib_hc_groups)) +
  geom_point(size = 6) +
  #The standard FI rate for green algae is 34, which is the average FI of the green alga in the control treatment acclimated to chlamy-only culture in 2025.
  geom_hline(yintercept = 34, color = 'grey30',linetype = 'dashed') +
  geom_vline(xintercept = 1, color = 'grey30',linetype = 'dotted') +
  #ggrepel::geom_text_repel(aes(label = Strain), max.overlaps = 20,
  #                           size = 2, alpha = 0.8,fontface = 'bold', show.legend = F) +
  scale_color_manual(
    values = c('plum4','thistle','goldenrod2'),
    name = paste0('Inhibition Groups',"\n",'(Hierarchical Clustering)')) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  labs(title = 'Defense traits of Microcystis WLECC strains',
       x = 'Mussel selectivity for Microcystis (Unpalatability)',
       y = 'Total amount of phytoplankton ingested by mussel (Inhibition)') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=16,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=16,color='black',face = "bold"),
        axis.title = element_text(size=10,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank())

ggsave('figures/fig2A_inhib_hc_groups.png', width = 6, height = 4, dpi = 300)


```

## DAPC on ANI variation

```{r}
# Prepare table
bgc_ani_wide_trait <- bgc_ani_wide %>%
  inner_join(defense_trait_DAPC, by = c('Strain' = 'Strain'))
```

```{r}
# initiate top loading list
top_loading_list <- list()
```

## ANI and palatability

```{r}
bgc_ani_no_inhibitory <- bgc_ani_wide_trait %>%
  filter(unpala_hc_groups != 'Not included')
gen_matrix <- as.matrix(bgc_ani_no_inhibitory[2:17])
palability_vector <- bgc_ani_no_inhibitory %>%
  pull(arrang_unpala_hc_groups) 

ani_dapc <- dapc(gen_matrix, palability_vector, n.da = 4, n.pca = 10)
scatter(ani_dapc, scree.da = TRUE, posi.da = 'bottomright', legend = TRUE, cex = 1.5)

# overlay loadings on the DAPC plot
loadingplot(ani_dapc$var.contr, axis = 1, thres = 0.2, lab.jitter = 1)
loadingplot(ani_dapc$var.contr, axis = 2, thres = 0.2, lab.jitter = 1)
loadingplot(ani_dapc$var.contr, axis = 3, thres = 0.2, lab.jitter = 1)
```

Evaluate the DAPC model using cross-validation to determine the optimal number of principal components to retain.

```{r}
set.seed(123)

ani_dapc_unp_xval <- xvalDapc(gen_matrix, palability_vector, n.pca.max = 20, training.set = 0.9,
                             result = "groupMean", center = TRUE, scale = FALSE,
                             n.rep = 30, xval.plot = TRUE)

```


```{r}
## ggplot version of DAPC plot
scores_ani_dapc <- as.data.frame(ani_dapc$ind.coord) 
scores_ani_dapc$groups <- palability_vector 
scores_ani_dapc$Strain <- bgc_ani_no_inhibitory$Strain

scaling_factor <- 5
loadings <- as.data.frame(ani_dapc$var.contr) %>%
  rownames_to_column(var = 'cluster') 
# filter the most contributing variables based on a threshold
top_loadings <- loadings %>%
  mutate(keep = case_when(
    LD1 > quantile(LD1, 0.85) | LD2 > quantile(LD2, 0.85)  ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  filter(keep == TRUE)

top_loading_list[['ani_unpala']] <- top_loadings
  

ggplot(scores_ani_dapc, aes(x = LD1, y = LD2,color = groups)) +
  geom_point(size = 6) +
  scale_color_manual(
    values = c('lightpink4','rosybrown1','darkseagreen3','aquamarine4',"grey70"),
    name = 'Unpalatability Groups') +
  #ggrepel::geom_text_repel(
  #  aes(label = Strain), max.overlaps = 20,
  #  size = 2, alpha = 1,fontface = 'bold') +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey50') +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'grey50') +
  labs(title = 'DAPC of BGC ANI with Unpalatability Groups',
       x = paste0('LD1 (', round(ani_dapc$eig[1]/sum(ani_dapc$eig)*100, 2), '%)'),
       y = paste0('LD2 (', round(ani_dapc$eig[2]/sum(ani_dapc$eig)*100, 2), '%)'),
       color = 'Unpalatability Groups') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=16,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank()) +
  geom_segment(data=top_loadings, aes(x=0, y=0, xend=LD1*scaling_factor, yend=LD2*scaling_factor), 
               arrow=arrow(length=unit(0.2,"cm")), color="black", alpha = 0.6) +
  ggrepel::geom_text_repel(data=top_loadings, aes(x=LD1*scaling_factor, y=LD2*scaling_factor, label=cluster), 
                  size=4, color="black", fontface = 'bold')
ggsave('figures/fig2A_ani_DAPC_unpala_plot.png', width = 6, height = 4, dpi = 300)
```

## ANI and clearance rate

```{r}
gen_matrix <- as.matrix(bgc_ani_wide_trait[2:17])
inhib_vector <- bgc_ani_wide_trait %>%
  pull(inhib_hc_groups)

ani_dapc <- dapc(gen_matrix, inhib_vector, n.da = 3, n.pca = 10)
scatter(ani_dapc, scree.da = TRUE, posi.da = 'bottomright', legend = TRUE, cex = 1.5)

# overlay loadings on the DAPC plot
loadingplot(ani_dapc$var.contr, axis = 1, thres = 0.2, lab.jitter = 1)
loadingplot(ani_dapc$var.contr, axis = 2, thres = 0.2, lab.jitter = 1)
```

Evaluate the DAPC model using cross-validation to determine the optimal number of principal components to retain.

```{r}
set.seed(123)

ani_dapc_inhib_xval <- xvalDapc(gen_matrix, inhib_vector, n.pca.max = 20, training.set = 0.9,
                             result = "groupMean", center = TRUE, scale = FALSE,
                             n.rep = 30, xval.plot = TRUE)


```

```{r}
## ggplot version of DAPC plot
scores_ani_dapc <- as.data.frame(ani_dapc$ind.coord) 
scores_ani_dapc$groups <- inhib_vector 
scores_ani_dapc$Strain <- bgc_ani_wide_trait$Strain

scaling_factor <- 5
loadings <- as.data.frame(ani_dapc$var.contr) %>%
  rownames_to_column(var = 'cluster') 
# filter the most contributing variables based on a threshold
top_loadings <- loadings %>%
  mutate(keep = case_when(
    LD1 > quantile(LD1, 0.85) | LD2 > quantile(LD2, 0.85)  ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  filter(keep == TRUE)

top_loading_list[['ani_inhib']] <- top_loadings
  

ggplot(scores_ani_dapc, aes(x = LD1, y = LD2,color = groups)) +
  geom_point(size = 6) +
  scale_color_manual(
    values = c('plum4','thistle','goldenrod2'),
    name = 'Inhibition Groups') +
  #ggrepel::geom_text_repel(
  #  aes(label = Strain), max.overlaps = 20,
  #  size = 2, alpha = 1,fontface = 'bold') +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey50') +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'grey50') +
  labs(title = 'DAPC of BGC ANI with Inhibition Groups',
       x = paste0('LD1 (', round(ani_dapc$eig[1]/sum(ani_dapc$eig)*100, 2), '%)'),
       y = paste0('LD2 (', round(ani_dapc$eig[2]/sum(ani_dapc$eig)*100, 2), '%)'),
       color = 'Inhibition Groups') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=16,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank()) +
  geom_segment(data=top_loadings, aes(x=0, y=0, xend=LD1*scaling_factor, yend=LD2*scaling_factor), 
               arrow=arrow(length=unit(0.2,"cm")), color="black", alpha = 0.6) +
  ggrepel::geom_text_repel(data=top_loadings, aes(x=LD1*scaling_factor, y=LD2*scaling_factor, label=cluster), 
                  size=4, color="black", fontface = 'bold')
ggsave('figures/fig2B_ani_DAPC_inhib_plot.png', width = 5.6, height = 4, dpi = 300)
```

## Gene content conservation and traits

```{r}
# Prepare table
bgc_pres_wide_trait <- bgc_pres_wide %>%
  inner_join(defense_trait_DAPC, by = c('Strain' = 'Strain'))
```


```{r}
gen_matrix <- as.matrix(bgc_pres_wide_trait[2:17])
inhib_vector <- bgc_pres_wide_trait %>%
  pull(inhib_hc_groups)

pres_dapc <- dapc(gen_matrix, inhib_vector, n.da = 3, n.pca = 10)
scatter(pres_dapc, scree.da = TRUE, posi.da = 'bottomright', legend = TRUE, cex = 1.5)

# overlay loadings on the DAPC plot
loadingplot(pres_dapc$var.contr, axis = 1, thres = 0.1, lab.jitter = 1)
loadingplot(pres_dapc$var.contr, axis = 2, thres = 0.1, lab.jitter = 1)
```

Evaluate the DAPC model using cross-validation to determine the optimal number of principal components to retain.

```{r}
set.seed(123)

pres_dapc_inhib_xval <- xvalDapc(gen_matrix, inhib_vector, n.pca.max = 20, training.set = 0.9,
                             result = "groupMean", center = TRUE, scale = FALSE,
                             n.rep = 30, xval.plot = TRUE)

```

```{r}
## ggplot version of DAPC plot
scores_pres_dapc <- as.data.frame(pres_dapc$ind.coord) 
scores_pres_dapc$groups <- inhib_vector 
scores_pres_dapc$Strain <- bgc_pres_wide_trait$Strain

scaling_factor <- 5
loadings <- as.data.frame(pres_dapc$var.contr) %>%
  rownames_to_column(var = 'cluster') 
# filter the most contributing variables based on a threshold
top_loadings <- loadings %>%
  mutate(keep = case_when(
    LD1 > quantile(LD1, 0.85) | LD2 > quantile(LD2, 0.85)  ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  filter(keep == TRUE)
  
top_loading_list[['pres_inhib']] <- top_loadings

ggplot(scores_pres_dapc, aes(x = LD1, y = LD2,color = groups)) +
  geom_point(size = 6) +
  scale_color_manual(
    values = c('plum4','thistle','goldenrod2'),
    name = 'Inhibition Groups') +
  #ggrepel::geom_text_repel(
  #  aes(label = Strain), max.overlaps = 20,
  #  size = 2, alpha = 1,fontface = 'bold') +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey50') +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'grey50') +
  labs(title = 'DAPC of BGC Perc pres with Inhibition Groups',
       x = paste0('LD1 (', round(pres_dapc$eig[1]/sum(pres_dapc$eig)*100, 2), '%)'),
       y = paste0('LD2 (', round(pres_dapc$eig[2]/sum(pres_dapc$eig)*100, 2), '%)'),
       color = 'Inhibition Group') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=16,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank()) +
  geom_segment(data=top_loadings, aes(x=0, y=0, xend=LD1*scaling_factor, yend=LD2*scaling_factor), 
               arrow=arrow(length=unit(0.2,"cm")), color="black", alpha = 0.6) +
  ggrepel::geom_text_repel(data=top_loadings, aes(x=LD1*scaling_factor, y=LD2*scaling_factor, label=cluster), 
                  size=4, color="black", fontface = 'bold')
ggsave('figures/fig2B_pres_DAPC_inhib_plot.png', width = 5.6, height = 4, dpi = 300)
```

```{r}
bgc_pres_no_inhibitory <- bgc_pres_wide_trait %>%
  filter(unpala_hc_groups != 'Not included')
gen_matrix <- as.matrix(bgc_pres_no_inhibitory[2:17])
palability_vector <- bgc_pres_no_inhibitory %>%
  pull(arrang_unpala_hc_groups) 


pres_dapc <- dapc(gen_matrix, palability_vector, n.da = 4, n.pca = 10)
scatter(pres_dapc, scree.da = TRUE, posi.da = 'bottomright', legend = TRUE, cex = 1.5)

# overlay loadings on the DAPC plot
loadingplot(pres_dapc$var.contr, axis = 1, thres = 0.2, lab.jitter = 1)
loadingplot(pres_dapc$var.contr, axis = 2, thres = 0.2, lab.jitter = 1)
loadingplot(pres_dapc$var.contr, axis = 3, thres = 0.2, lab.jitter = 1)
```

Evaluate the DAPC model using cross-validation to determine the optimal number of principal components to retain.

```{r}
set.seed(123)

pres_dapc_unp_xval <- xvalDapc(gen_matrix, palability_vector, n.pca.max = 10, training.set = 0.9,
                             result = "groupMean", center = TRUE, scale = FALSE,
                             n.rep = 30, xval.plot = TRUE)

```

```{r}
## ggplot version of DAPC plot
scores_pres_dapc <- as.data.frame(pres_dapc$ind.coord) 
scores_pres_dapc$groups <- palability_vector 
scores_pres_dapc$Strain <- bgc_pres_no_inhibitory$Strain

scaling_factor <- 5
loadings <- as.data.frame(pres_dapc$var.contr) %>%
  rownames_to_column(var = 'cluster') 
# filter the most contributing variables based on a threshold
top_loadings <- loadings %>%
  mutate(keep = case_when(
    LD1 > quantile(LD1, 0.85) | LD2 > quantile(LD2, 0.85)  ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  filter(keep == TRUE)
  
top_loading_list[['pres_unpala']] <- top_loadings

ggplot(scores_pres_dapc, aes(x = LD1, y = LD2,color = groups)) +
  geom_point(size = 6) +
  scale_color_manual(
    values = c('lightpink4','rosybrown1','aquamarine4','darkseagreen3'),
    name = 'Unpalatability Groups') +
  #ggrepel::geom_text_repel(
  #  aes(label = Strain), max.overlaps = 20,
  #  size = 2, alpha = 1,fontface = 'bold') +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey50') +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'grey50') +
  labs(title = 'DAPC of BGC Perc pres with Palatability Groups',
       x = paste0('LD1 (', round(ani_dapc$eig[1]/sum(ani_dapc$eig)*100, 2), '%)'),
       y = paste0('LD2 (', round(ani_dapc$eig[2]/sum(ani_dapc$eig)*100, 2), '%)'),
       color = 'Unpalatability Groups') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(linewidth=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=16,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank()) +
  geom_segment(data=top_loadings, aes(x=0, y=0, xend=LD1*scaling_factor, yend=LD2*scaling_factor), 
               arrow=arrow(length=unit(0.2,"cm")), color="black", alpha = 0.6) +
  ggrepel::geom_text_repel(data=top_loadings, aes(x=LD1*scaling_factor, y=LD2*scaling_factor, label=cluster), 
                  size=4, color="black", fontface = 'bold')
ggsave('figures/fig2A_pres_DAPC_unpala_plot.png', width = 6, height = 4, dpi = 300)
```

## Variance explained by N-rich or C-rich BGCs

```{r}
## Input N- or C-rich annotation
bgc_nc_richanno <- read.csv('inputs/filtered_bgc_variation_summary_LH.csv', header = T) %>%
  select(cluster, Type, Type_2)
## Genetic variants (gene content conservation) and defense traits
bgc_pres_nc_cat <- bgc_trait_dbclean %>%
  filter(!cluster %in% low_present_cluster) %>%
  left_join(bgc_nc_richanno, by = 'cluster') %>%
  select(Strain, cluster, Percent_presence, Type, Type_2) %>%
  inner_join(defense_trait_DAPC, by = c('Strain' = 'Strain')) %>%
  group_by(Strain, Type) %>%
  summarise(Percent_presence = mean(Percent_presence), 
            Unpalatability = mean(Unpalatability), 
            Inhibition = mean(Inhibition),.groups = 'drop') %>%
  pivot_longer(cols = c('Unpalatability','Inhibition'), names_to = 'Trait', values_to = 'Trait_value') 

## plot gene content conservation variation with defense traits
ggplot(bgc_pres_nc_cat, aes(x = Percent_presence, y = Trait_value, color = Type)) +
  geom_point(size = 3) +
  geom_smooth(method = 'lm', se = TRUE, color = 'black', linetype = 'dashed') +
  scale_color_manual(values = c('cyan4','indianred4'), name = 'BGC Type') +
  facet_wrap(Trait~Type, scale = "free_y") +
  labs(title = NULL,
       x = 'Gene content conservation (%)',
       y = 'Trait Values') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=12,color='black',face = "bold"),
        strip.text = element_text(size = 12, color = 'black',face = "bold"),
        strip.background = element_rect(fill = "grey100", 
                                        color = NA, linewidth = 1),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=12,color='black',hjust=0, face = "bold"),
        legend.text = element_text(size=10,color='black',hjust=0),
        legend.position = 'right',
        panel.grid.major = element_blank())

```

```{r}
## linear regression models
inh_C_pres_lm <- lm(Trait_value ~ Percent_presence, data = bgc_pres_nc_cat %>% filter(Type == 'C-Rich') %>% filter(Trait == 'Inhibition'))

inh_N_pres_lm <- lm(Trait_value ~ Percent_presence, data = bgc_pres_nc_cat %>% filter(Type == 'N-Rich') %>% filter(Trait == 'Inhibition'))

unp_C_pres_lm <- lm(Trait_value ~ Percent_presence, data = bgc_pres_nc_cat %>% filter(Type == 'C-Rich') %>% filter(Trait == 'Unpalatability'))

unp_N_pres_lm <- lm(Trait_value ~ Percent_presence, data = bgc_pres_nc_cat %>% filter(Type == 'N-Rich') %>% filter(Trait == 'Unpalatability'))

shapiro.test(inh_C_pres_lm$residuals);shapiro.test(inh_N_pres_lm$residuals);shapiro.test(unp_C_pres_lm$residuals);shapiro.test(unp_N_pres_lm$residuals)

summary(inh_C_pres_lm);summary(inh_N_pres_lm);summary(unp_C_pres_lm);summary(unp_N_pres_lm)
```

```{r}
## Genetic variants (shared gene ANI) and defense traits
bgc_ani_nc_cat <- bgc_trait_dbclean %>%
  filter(!cluster %in% low_present_cluster) %>%
  select(Strain, cluster, ANI) %>%
  left_join(bgc_nc_richanno, by = 'cluster') %>%
  inner_join(defense_trait_DAPC, by = c('Strain' = 'Strain')) %>%
  group_by(Strain, Type) %>%
  summarise(ANI = mean(ANI), 
            Unpalatability = mean(Unpalatability), 
            Inhibition = mean(Inhibition),.groups = 'drop') %>%
  pivot_longer(cols = c('Unpalatability','Inhibition'), names_to = 'Trait', values_to = 'Trait_value')

## plot shared gene ANI variation with defense traits
ggplot(bgc_ani_nc_cat, aes(x = ANI, y = Trait_value, color = Type)) +
  geom_point(size = 3) +
  geom_smooth(method = 'lm', se = TRUE, color = 'black', linetype = 'dashed') +
  scale_color_manual(values = c('cyan4','indianred4'), name = 'BGC Type') +
  facet_wrap(Trait~Type, scale = "free_y") +
  labs(title = NULL,
       x = 'Shared gene ANI (%)',
       y = 'Trait Values') +
  theme_pubclean()+
  theme(plot.title= element_text(size=12,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(hjust = 0.5,size=10,color='black',face = "bold"),
        axis.title = element_text(size=12,color='black',face = "bold"),
        strip.text = element_text(size = 12, color = 'black',face = "bold"),
        strip.background = element_rect(fill = "grey100", 
                                        color = NA, linewidth = 1),
        panel.background = element_rect(fill = "grey100"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=12,color='black',hjust=0, face = "bold"),
        legend.text = element_text(size=10,color='black',hjust=0),
        legend.position = 'right',
        panel.grid.major = element_blank())
```

```{r}
## linear regression models
inh_C_ani_lm <- lm(Trait_value ~ ANI, data = bgc_ani_nc_cat %>% filter(Type == 'C-Rich') %>% filter(Trait == 'Inhibition'))
inh_N_ani_lm <- lm(Trait_value ~ ANI, data = bgc_ani_nc_cat %>% filter(Type == 'N-Rich') %>% filter(Trait == 'Inhibition'))
unp_C_ani_lm <- lm(Trait_value ~ ANI, data = bgc_ani_nc_cat %>% filter(Type == 'C-Rich') %>% filter(Trait == 'Unpalatability'))
unp_N_ani_lm <- lm(Trait_value ~ ANI, data = bgc_ani_nc_cat %>% filter(Type == 'N-Rich') %>% filter(Trait == 'Unpalatability'))

shapiro.test(inh_C_ani_lm$residuals);shapiro.test(inh_N_ani_lm$residuals);shapiro.test(unp_C_ani_lm$residuals);shapiro.test(unp_N_ani_lm$residuals)

summary(inh_C_ani_lm);summary(inh_N_ani_lm);summary(unp_C_ani_lm);summary(unp_N_ani_lm)
```
## BGC correlations

```{r}
# Prepare top loadings table
top_loadings_table <- bind_rows(top_loading_list, .id = 'DAPC_analysis') %>%
  select(cluster, DAPC_analysis)
```


```{r}
# correlation matrix for gene content conservation
## Calculate the correlation matrix 
bgc_pres_matrix <- bgc_pres_wide %>%
  column_to_rownames(var = 'Strain') %>%
  as.matrix() 

## Calculate the correlation matrix and p-values
bgc_pres_cor <- Hmisc::rcorr(bgc_pres_matrix, type = 'pearson')


# plot correlation matrix
ggcorrplot::ggcorrplot(bgc_pres_cor$r, 
                       hc.order = TRUE, 
                       hc.method = "average",
                       type = "upper", 
                       show.diag  = TRUE,
                       colors = c("cyan4", "white", "indianred4"),
                       #ggtheme = ggpubr::theme_pubclean(),
                       title = "Correlation Matrix of BGC variation (gene content conservation)") +
  theme(plot.title= element_text(size=12,color='black',hjust=0.5,face = "bold"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(size=10,color='black',face = "bold",angle = 90),
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
ggsave('figures/supplFig2A_bgc_pres_corr.png', width = 6, height = 6, dpi = 300)
```

```{r} 
# correlation among BGCs based on ANI
bgc_ani_matrix <- bgc_ani_wide %>%
  column_to_rownames(var = 'Strain') %>%
  as.matrix() 

## Calculate the correlation matrix and p-values
bgc_ani_cor <- Hmisc::rcorr(bgc_ani_matrix, type = 'pearson')

# plot correlation matrix
ggcorrplot::ggcorrplot(bgc_ani_cor$r, 
                       hc.order = TRUE, 
                       hc.method = "average",
                       type = "upper", 
                       show.diag = TRUE,
                       colors = c("cyan4", "white", "indianred4"),
                       #ggtheme = ggpubr::theme_pubclean(),
                       title = "Correlation Matrix of BGC variation (shared gene ANI)") +
  theme(plot.title= element_text(size=12,color='black',hjust=0.5,face = "bold"),
        axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
        axis.text.x = element_text(size=10,color='black',face = "bold",angle = 90),
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.position = 'right',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
ggsave('figures/supplFig2A_bgc_ani_corr.png', width = 6, height = 6, dpi = 300)
```

