---
title: "ND_LEmetaG_lgt"
author: "Yu-Cheng Lin"
date: "2026-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(tidyverse)
library(vegan)
```

Load lgt abundance data in Nikesh's metagenome samples from the Lake Erie mussel feeding experiment from 2018-2019. The reads are mapped to Microcystis marker gene lgt_516 competitively with a identity cutoff at 99% and coverage cutoff at 50%.

```{r}
# Load lgt abundance data
lgt_abund <- read.csv("inputs/ND_samples_lgt_full_id99.mapped.counts.csv") 
clade_info <- read.table("inputs/microcytis.genotypes.txt", header = TRUE, sep = "\t") 
# Row bind the two data frames
lgt_abund <- lgt_abund %>%
  left_join(clade_info, by = "seqnames") %>%
  mutate(Clade = ifelse(is.na(Clade), "Unknown", Clade)) %>%
  unglue::unglue_unnest(Date, "{Year}_{Month}_{Day}", remove = F) 
# Give replicate labels to each sample
#rep_labels <- lgt_abund %>%
#  group_by(Year, Month, Date, lake,Treatment,SampleID) %>%
#  summarise(count = n()) 
#write.csv(rep_labels, "outputs/lgt_abundance_sample_replicates.csv", row.names = FALSE)
# Add rep labels to the lgt abundance data
rep_labels <- read.csv("inputs/lgt_abundance_sample_replicates.csv") 
reps <- rep_labels[,c("SampleID","Rep")]
lgt_abund <- lgt_abund %>%
  left_join(reps, by = "SampleID") %>%
  mutate(sample_name = paste0(Date, "_", Rep)) 
# Check NA values in the lgt abundance data
na_values <- lgt_abund %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "na_count")
na_samples <- lgt_abund %>%
  filter(is.na(TPM_relative_abundance)) 
lgt_abund <- lgt_abund %>%
  drop_na()
# Check if the relative abundance of a sample sums up to 1
rel_abund_check <- lgt_abund %>%
  group_by(SampleID) %>%
  summarise(total_rel_abund = sum(TPM_relative_abundance, na.rm = T)) %>%
  ungroup()
```

Check the distribution of relative abundance of lgt_516 across samples

```{r}
# RPKM
rpkm_plot <- ggplot(lgt_abund %>% filter(rpkm!=0), aes(x = Date, y= rpkm, fill=Treatment)) +
  geom_violin(aes(color = Treatment))+
  facet_wrap(~Year*lake, ncol = 3, scales = "free_x") +
  labs(title = "RPKM of lgt_516 marker genes in Mussel Feeding Experiment Samples",
       y = "RPKM", x = "Treatment") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.position = "right",
          panel.grid.major = element_blank())
## TPKM 
tpm_plot <- ggplot(lgt_abund %>% filter(TPM!=0), aes(x = Date, y= TPM, fill=Treatment)) +
  geom_violin(aes(color = Treatment))+
  facet_wrap(~Year*lake, ncol = 3, scales = "free_x") +
  labs(title = "TPM of lgt_516 marker genes in Mussel Feeding Experiment Samples",
       y = "TPM", x = "Treatment") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.position = "right",
          panel.grid.major = element_blank())
rpkm_plot;tpm_plot

```

# Relative abundance plot

Focus on TPM from this point forward. Plot relative abundance of lgt_516 across samples, arranged by sampling date and treatment groups

```{r}
# TPM
LE_plot_tpm <- ggplot(lgt_abund %>% filter(lake == "LE"),
                  aes(x = TPM_relative_abundance, y = sample_name, fill = seqnames)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~Treatment*Year, nrow = 2, scales = "free") +
  labs(title = "Microcystis strain dynamics in Lake Erie Quagga Mussel Feeding Experiments",
       y = "Sampling Date", x = "Relative Abundance (TPM)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "none",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/microcystis_lgt_tpm_relative_abundance_LE.png", width = 10, height = 6, dpi = 300)
plot_2019_TPM <- ggplot(lgt_abund %>% 
                          filter( Year == 19) %>% 
                          mutate(sample_name_trt = paste0(Treatment, "_", sample_name)), 
                    aes(x = TPM_relative_abundance, y = sample_name_trt, fill = seqnames)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~lake*Date, nrow = 2, scales = "free") +
  labs(title = "Microcystis strain dynamics in 2019 Quagga Mussel Feeding Experiments",
       y = "Sampling Date", x = "Relative Abundance") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.position = "none",
          panel.grid.major = element_blank())
#ggsave("figures/fig3C/microcystis_lgt_tpm_relative_abundance_2019.png", width = 10, height = 6, dpi = 300)
#legend_plot_tpm <- get_legend(LE_plot_tpm + theme(legend.position = "bottom"))
#grid::grid.newpage()
#grid::grid.draw(legend_plot_tpm)
#ggsave("figures/fig3C/microcystis_lgt_tpm_relative_abundance_LE_legend.png", width = 25, height = 20, dpi = 300)
LE_plot_tpm;plot_2019_TPM
```

Clade level dynamics

Summarize the relative abundance of lgt_516 by clade and plot the dynamics across samples

```{r}
clade_lgt_abund <- lgt_abund %>%
  group_by(Clade, SampleID, Date, Rep, Month, Year, lake, Treatment, sample_name) %>%
  summarise(sum_TPM_rel_abund = sum(TPM_relative_abundance, na.rm = T),
            sum_TPM = sum(TPM, na.rm = T)) %>%
  ungroup()
write.csv(clade_lgt_abund, "inputs/clade_lgt_abundance.csv", row.names = FALSE)
```

Plot TPM relative abundance

```{r}
LE2018_clade_plot_TPMrel <- ggplot(clade_lgt_abund %>% 
                          filter(lake == "LE" & Year == 18) %>% 
                          mutate(sample_name_trt = paste0(Treatment, "_", sample_name)),
                  aes(x = sum_TPM_rel_abund, y = sample_name_trt, fill = Clade)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~Date, nrow = 2, scales = "free") +
  labs(title = "Microcystis Clade Dynamics in 2018 Lake Erie Mussel Feeding Experiments",
       y = "Sampling Date", x = "Relative Abundance (TPM)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/microcystis_lgt_clade_tpm_relative_abundance_2018LE.png", width = 13, height = 6.5, dpi = 300)
LE2019_clade_plot_TPMrel <- ggplot(clade_lgt_abund %>% 
                          filter(lake == "LE" & Year == 19) %>% 
                          mutate(sample_name_trt = paste0(Treatment, "_", sample_name)),
                  aes(x = sum_TPM_rel_abund, y = sample_name_trt, fill = Clade)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~Date*Year, nrow = 2, scales = "free") +
  labs(title = "Microcystis Clade Dynamics in 2019 Lake Erie Mussel Feeding Experiments",
       y = "Sampling Date", x = "Relative Abundance (TPM)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/microcystis_lgt_clade_tpm_relative_abundance_2019LE.png", width = 9, height = 6, dpi = 300)
LH2019_clade_plot_TPMrel <- ggplot(clade_lgt_abund %>% 
                          filter(lake == "LH" & Year == 19) %>% 
                          mutate(sample_name_trt = paste0(Treatment, "_", sample_name)),
                  aes(x = sum_TPM_rel_abund, y = sample_name_trt, fill = Clade)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~Date*Year, nrow = 2, scales = "free") +
  labs(title = "Microcystis Clade Dynamics in 2019 Lake Huron Mussel Feeding Experiments",
       y = "Sampling Date", x = "Relative Abundance (TPM)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/microcystis_lgt_clade_tpm_relative_abundance_2019LH.png", width = 6, height = 6, dpi = 300)

LE2018_clade_plot_TPMrel;LE2019_clade_plot_TPMrel;LH2019_clade_plot_TPMrel


```

Just TPM

```{r}
LE2018_clade_plot_TPM <- ggplot(clade_lgt_abund %>% 
                          filter(lake == "LE" & Year == 18) %>% 
                          mutate(sample_name_trt = paste0(Treatment, "_", sample_name)),
                  aes(x = sum_TPM, y = sample_name_trt, fill = Clade)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~Date, nrow = 2, scales = "free") +
  labs(title = "Microcystis Clade Dynamics in 2018 Lake Erie Mussel Feeding Experiments",
       y = "Sampling Date", x = "Abundance (TPM)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/microcystis_lgt_clade_tpm_abundance_2018LE.png", width = 13, height = 6.5, dpi = 300)
LE2019_clade_plot_TPM <- ggplot(clade_lgt_abund %>% 
                          filter(lake == "LE" & Year == 19) %>% 
                          mutate(sample_name_trt = paste0(Treatment, "_", sample_name)),
                  aes(x = sum_TPM, y = sample_name_trt, fill = Clade)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~Date*Year, nrow = 2, scales = "free") +
  labs(title = "Microcystis Clade Dynamics in 2019 Lake Erie Mussel Feeding Experiments",
       y = "Sampling Date", x = "Abundance (TPM)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/microcystis_lgt_clade_tpm_abundance_2019LE.png", width = 9, height = 6, dpi = 300)
LH2019_clade_plot_TPM <- ggplot(clade_lgt_abund %>% 
                          filter(lake == "LH" & Year == 19) %>% 
                          mutate(sample_name_trt = paste0(Treatment, "_", sample_name)),
                  aes(x = sum_TPM, y = sample_name_trt, fill = Clade)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~Date*Year, nrow = 2, scales = "free") +
  labs(title = "Microcystis Clade Dynamics in 2019 Lake Huron Mussel Feeding Experiments",
       y = "Sampling Date", x = "Abundance (TPM)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/microcystis_lgt_clade_tpm_abundance_2019LH.png", width = 6, height = 6, dpi = 300)

LE2018_clade_plot_TPM;LE2019_clade_plot_TPM;LH2019_clade_plot_TPM


```

# Clearance rate and strain abundance

```{r}
## load chlorophyll A data
chlA <- read.csv("inputs/fp_data/ND_2019_chla.csv") %>%
  unglue::unglue_unnest(ExDate, "{month}/{day}/{year}", remove = F) %>%
  mutate(Date = paste0(year, "_0", month, "_", day)) %>%
  mutate(chla_diff = I_CHL-F_CHL)  %>%
  rename(lake = Lake) %>%
  mutate(lake = case_when(
    lake == "LE" ~ "LE",
    lake == "SB" ~ "LH"
  )) 
ctrl_chlA_diff <- chlA %>%
  filter(Treatment == "control") %>%
  group_by(Date, lake) %>%
  summarise(mean_ctrl_diff = mean(chla_diff, na.rm = T),
            mean_F_CHL = mean(F_CHL, na.rm = T)) %>%
  ungroup()
perc_consumed_chlA <- chlA %>%
  filter(Treatment == "quagga") %>%
  select(Date, lake, Rep, Treatment, chla_diff) %>%
  left_join(ctrl_chlA_diff, by = c("Date", "lake")) %>%
  mutate(perc_consumed = (chla_diff - mean_ctrl_diff)/mean_F_CHL)
## Combine control abundance with clearance rate
lgt_clearance <- clade_lgt_abund %>%
  filter(Treatment == "control" & Year != 18) %>%
  left_join(perc_consumed_chlA %>% select(Date, lake, Rep, perc_consumed), by = c("Date", "Rep", "lake"))
```


```{r}
## Select abundant clades
filter_clade <- lgt_clearance %>%
  group_by(Clade) %>%
  summarise(max_rel_abund = max(sum_TPM_rel_abund, na.rm = T)) %>%
  ungroup() %>%
  filter(max_rel_abund > 10) %>%# third quantile
  pull(Clade)
## Plot clade to clearance rate
for (clade in filter_clade) {
  print(clade)
  p <- ggplot(lgt_clearance %>% filter(Clade == clade), aes(x = sum_TPM_rel_abund, y = perc_consumed)) +
    geom_point(aes(color = Date), size = 3) +
    facet_wrap(~lake, scales = "fixed") +
    labs(title = paste0("Relative Abundance of ", clade, " Clade vs Clearance Rate"),
         x = paste0("Relative Abundance of ", clade, " Clade in Control"), y = "Percentage of Chlorophyll A Consumed") +
    theme_pubclean()+
      theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
            axis.line = element_line(linewidth=0, colour = "black"),
            axis.ticks = element_line(size=1, colour = "black"),
            axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
            axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
            axis.title = element_text(size=15,color='black',face = "bold"),
            panel.background = element_rect(fill = "grey100"),
            panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
            legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
            legend.text = element_text(size = 8),
            legend.box = "vertical",
            legend.position = "right",
            panel.grid.major = element_blank())
  print(p)
  #ggsave(p, filename = paste0("figures/fig3C/", gsub('/','|',clade), "_clearance_rate.png"), width = 6, height = 5, dpi = 300)
}
```


## Strain composition and clearance rate

Calculate bray-curtis distance of the control samples and plot the distance with clearance rate

```{r}
strain_comp <- lgt_abund %>%
  filter(Treatment == "control") %>%
  group_by(seqnames, Date) %>%
  summarize(avg_TPM_rel_abund = mean(TPM_relative_abundance)) %>%
  pivot_wider(names_from = seqnames, values_from = avg_TPM_rel_abund) %>%
  column_to_rownames(var = "Date") %>%
  mutate(check_sum_comp = rowSums(across(where(is.numeric)), na.rm = TRUE))
         
strain_comp_dist <- vegdist(strain_comp, method = "bray")
```

```{r}
strain_pcoa <- cmdscale(strain_comp_dist, eig = TRUE)
var_exp <- pmax(strain_pcoa$eig, 0) / sum(pmax(strain_pcoa$eig, 0))

strain_pcoa_scores <- as.data.frame(strain_pcoa$points) %>%
  rownames_to_column(var = "Date") %>%
  left_join(lgt_abund %>%
              select(Date, lake, Month, Year, Day) %>% 
              distinct(), by = "Date") %>%
  mutate(Year = paste0(20,Year)) %>%
  mutate(lake = case_when(
    lake == "LE" ~ "Lake Erie",
    lake == "LH" ~ "Lake Huron"
  )) %>%
  mutate(sample_date = paste(Month, Day, sep = "/")) %>%
  mutate(Year_lake = paste(Year,lake, sep = " ")) 

ggplot(strain_pcoa_scores, aes(x = V1, y = V2)) +
  geom_point(aes(fill = Year_lake, shape = Month),size = 4) +
  scale_fill_manual(name = "Lake System", 
                    values = c("2018 Lake Erie" = "peachpuff2", 
                               "2019 Lake Erie" = "cyan4", 
                               "2019 Lake Huron" = "lightpink4")) +
  scale_shape_manual(name = "Month", 
                     values = c("06" = 21, "07" = 22, 
                                "08" = 23,"09" = 24, "10" = 25))+
  guides(fill = guide_legend(override.aes = list(shape = 21)),
         alpha = "none",
         shape = guide_legend(override.aes = list(fill = "grey80"))) +
  labs(title = "PCoA of Strain Composition in Control Samples",
       x = paste0("PCo1(", round(var_exp[1]*100,1), "%)"), 
       y = paste0("PCo2(", round(var_exp[2]*100,1), "%)")) +
  ggrepel::geom_text_repel(aes(label = sample_date), size = 4, 
                           max.overlaps = 10, color = "black",
                           fontface = "bold") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())
ggsave("figures/fig3C/strain_composition_pcoa.png", width = 6.5, height = 5, dpi = 300)

```


```{r}
plot_var_exp <- data.frame(PCo =  1:length(var_exp), var_exp = var_exp)
## scree plot
ggplot(plot_var_exp, aes(x = PCo, y = var_exp)) +
  geom_bar(stat = 'identity', fill = 'grey30') +
  geom_line() +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = paste0(round(var_exp*100,1), "%")), size = 5, fontface = "bold", max.overlaps = 10) +
  labs(title = "Scree Plot of PCoA",
       x = "Principal Coordinates", y = "Variance Explained (%)") +
  theme_pubclean()+
    theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=10,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=12,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())

ggsave("figures/fig3C/strain_composition_PCoA_scree_plot.png", width = 6, height = 5, dpi = 300)

```



# Differential abundance analysis

## Strain dynamics

```{r}
library(edgeR)
library(limma)

## Prepare metadata and count matrix
metadata <- lgt_abund %>%
  select(SampleID, Treatment, Date) %>%
  distinct() %>%
  arrange(SampleID)
count_matrix <- lgt_abund %>%
  select(SampleID, seqnames,TPM_relative_abundance) %>%
  pivot_wider(names_from = SampleID, values_from = TPM_relative_abundance) %>%
  #log transform the count matrix
  column_to_rownames(var = "seqnames") %>%
  as.matrix() 
log_count_matrix <- log2(count_matrix + 1) # add pseudocount to avoid log(0)

## factor level
factors <- factor(paste0(metadata$Treatment,"_",metadata$Date))

## Create the design matrix
design <- model.matrix(~0 + factors)
colnames(design) <- levels(factors)

## contrasts
contrast_matrix <- makeContrasts(
  logqc_LE_18_06_28 = quagga_18_06_28 - control_18_06_28,
  logqc_LE_18_07_10 = quagga_18_07_10 - control_18_07_10,
  logqc_LE_18_07_25 = quagga_18_07_25 - control_18_07_25,
  logqc_LE_18_08_15 = quagga_18_08_15 - control_18_08_15,
  logqc_LE_18_10_4 = quagga_18_10_4 - control_18_10_4,
  logqc_LH_19_06_12 = quagga_19_06_12 - control_19_06_12,
  logqc_LE_19_07_18 = quagga_19_07_18 - control_19_07_18,
  logqc_LE_19_07_31 = quagga_19_07_31 - control_19_07_31,
  logqc_LE_19_08_13 = quagga_19_08_13 - control_19_08_13,
  logqc_LH_19_08_22 = quagga_19_08_22 - control_19_08_22,
  logqc_LE_19_09_18 = quagga_19_09_18 - control_19_09_18,
  
  levels = design
)
```

```{r}
# fit linear model
fit <- lmFit(log_count_matrix, design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, trend = TRUE)

# extract results for each contrast
results_list <- list()
for (contrast in colnames(contrast_matrix)) {
  contrast_tb <- topTable(fit2, coef = contrast, number = Inf, adjust.method = "BH")
  filt_contrast_tb <- contrast_tb %>%
    rownames_to_column(var = "seqnames") %>%
    mutate(significant = case_when(
      adj.P.Val < 0.05 & logFC > 0 ~ "Increased",
      adj.P.Val < 0.05 & logFC < -0 ~ "Decreased",
      TRUE ~ "Not Significant"
    )) %>%
    arrange(adj.P.Val) %>%
    mutate(contrast = contrast) %>%
    mutate(Year = str_extract(contrast, "18|19"),
           lake = str_extract(contrast, "LE|LH"),
           Date = str_extract(contrast, "\\d{2}_\\d{2}_\\d{1,2}"))%>%
  separate(Date, into = c("Year", "Month", "Day"), sep = "_", remove = F, extra = "merge") 
  results_list[[contrast]] <- filt_contrast_tb
  
}

results_df <- bind_rows(results_list) %>%
  left_join(clade_info, by = "seqnames") %>%
  mutate(Clade = case_when(
    is.na(Clade) ~ "Unclassified",
    Clade == "Un" ~ "Unclassified",
    TRUE ~ Clade
  )) %>%
  mutate(Year = paste0(20,Year)) %>%
  mutate(lake = case_when(
    lake == "LE" ~ "Lake Erie",
    lake == "LH" ~ "Lake Huron"
  )) %>%
  mutate(sample_date = paste(Month, Day, sep = "/")) %>%
  mutate(Year_lake = paste(Year, lake, sep = " ")) %>%
  mutate(sig_label = case_when(
      significant != "Not Significant" ~ Year_lake,
      significant == "Not Significant" ~ "No change",
    )) %>%
  mutate(seqnames = str_replace(seqnames, "Microcystis_aeruginosa_", "")) %>%
  mutate(seqnames = str_replace(seqnames, "Microcystis_flos-aquae_", "")) %>%
  mutate(seqnames = str_replace(seqnames, "Microcystis_novacekii_", "")) %>%
  mutate(seqnames = str_replace(seqnames, "Microcystis_viridis_", "")) %>%
  mutate(seqnames = str_replace(seqnames, "Microcystis_panniformis_", "")) %>%
  mutate(seqnames = str_replace(seqnames, "Microcystis_wesenbergii_", "")) %>%
  mutate(seqnames = str_replace(seqnames, "Microcystis_", "")) %>%
  # remove anything after ";"
  mutate(seqnames = str_replace(seqnames, ";.*", "")) %>%
  mutate(strain_clade = paste0(seqnames, " (",Clade,")"))

write.csv(results_df, "output_tables/Microcystis_strain_differential_abundance_results.csv", row.names = FALSE)
```

Volcano plots

```{r}
ggplot(results_df, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(shape = Month, fill = sig_label, alpha = sig_label),color = "black",stroke = 0.3, size = 4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  #ggrepel::geom_text_repel(data = results_df %>% 
  #                           filter(adj.P.Val < 0.05 & abs(logFC) > 0.1), 
  #                         aes(label = strain_clade), size = 2, max.overlaps = 8, fontface = "bold") +
  #facet_grid(cols = vars(Year_lake), scales = "fixed") +
  scale_alpha_manual(values = c("2018 Lake Erie" = 1, 
                                "2019 Lake Erie" = 1, 
                                "2019 Lake Huron" = 1,"No change" = 0.5))+
  scale_fill_manual(name = "Lake System", 
                    values = c("2018 Lake Erie" = "peachpuff2", 
                               "2019 Lake Erie" = "cyan4", 
                               "2019 Lake Huron" = "lightpink4",
                               "No change" = "grey80")) +
  scale_shape_manual(name = "Month", 
                     values = c("06" = 21, "07" = 22, 
                                "08" = 23,"09" = 24, "10" = 25))+
  guides(fill = guide_legend(override.aes = list(shape = 21)),
         alpha = "none",
         shape = guide_legend(override.aes = list(fill = "grey80"))) +
  labs(title = "Hierarchy of vulnerability within Microcystis",
       x = "Log2 Fold Change (Quagga vs Control)", y = "-Log10 Adjusted P-value") +
  theme_pubclean()+
  theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())

ggsave("figures/fig3C/microcystis_log2QC_lgt_relabund_volcano_year_lake.png", width = 6, height = 4.5, dpi = 300)

```

## Clade dynamics

```{r}
## Prepare metadata and count matrix
metadata <- clade_lgt_abund %>%
  select(SampleID, Treatment, Date) %>%
  distinct() %>%
  arrange(SampleID)
count_matrix <- clade_lgt_abund %>%
  select(SampleID, Clade,sum_TPM_rel_abund) %>%
  pivot_wider(names_from = SampleID, values_from = sum_TPM_rel_abund) %>%
  #log transform the count matrix
  column_to_rownames(var = "Clade") %>%
  as.matrix() 
log_count_matrix <- log2(count_matrix + 1) # add pseudocount to avoid log(0)

## factor level
factors <- factor(paste0(metadata$Treatment,"_",metadata$Date))

## Create the design matrix
design <- model.matrix(~0 + factors)
colnames(design) <- levels(factors)

## contrasts
contrast_matrix <- makeContrasts(
  logqc_LE_18_06_28 = quagga_18_06_28 - control_18_06_28,
  logqc_LE_18_07_10 = quagga_18_07_10 - control_18_07_10,
  logqc_LE_18_07_25 = quagga_18_07_25 - control_18_07_25,
  logqc_LE_18_08_15 = quagga_18_08_15 - control_18_08_15,
  logqc_LE_18_10_4 = quagga_18_10_4 - control_18_10_4,
  logqc_LH_19_06_12 = quagga_19_06_12 - control_19_06_12,
  logqc_LE_19_07_18 = quagga_19_07_18 - control_19_07_18,
  logqc_LE_19_07_31 = quagga_19_07_31 - control_19_07_31,
  logqc_LE_19_08_13 = quagga_19_08_13 - control_19_08_13,
  logqc_LH_19_08_22 = quagga_19_08_22 - control_19_08_22,
  logqc_LE_19_09_18 = quagga_19_09_18 - control_19_09_18,
  
  levels = design
)
```

```{r}
# fit linear model
fit <- lmFit(log_count_matrix, design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, trend = TRUE)

# extract results for each contrast
results_list <- list()
for (contrast in colnames(contrast_matrix)) {
  contrast_tb <- topTable(fit2, coef = contrast, number = Inf, adjust.method = "BH")
  filt_contrast_tb <- contrast_tb %>%
    #rownames_to_column(var = "seqnames") %>%
    rownames_to_column(var = "Clade") %>%
    mutate(significant = case_when(
      adj.P.Val < 0.05 & logFC > 0 ~ "Increased",
      adj.P.Val < 0.05 & logFC < -0 ~ "Decreased",
      TRUE ~ "Not Significant"
    )) %>%
    arrange(adj.P.Val) %>%
    mutate(contrast = contrast) %>%
    mutate(Year = str_extract(contrast, "18|19"),
           lake = str_extract(contrast, "LE|LH"),
           Date = str_extract(contrast, "\\d{2}_\\d{2}_\\d{1,2}"))%>%
  separate(Date, into = c("Year", "Month", "Day"), sep = "_", remove = F, extra = "merge") 
  results_list[[contrast]] <- filt_contrast_tb
  
}

clade_results_df <- bind_rows(results_list) %>%
  mutate(Clade = case_when(
    is.na(Clade) ~ "Unknown",
    Clade == "Un" ~ "Unknown",
    TRUE ~ Clade
  )) %>%
  mutate(Year = paste0(20,Year)) %>%
  mutate(lake = case_when(
    lake == "LE" ~ "Lake Erie",
    lake == "LH" ~ "Lake Huron"
  )) %>%
  mutate(sample_date = paste(Month, Day, sep = "/")) %>%
  mutate(Year_lake = paste(Year, lake, sep = " ")) %>%
  mutate(sig_label = case_when(
      significant != "Not Significant" ~ Year_lake,
      significant == "Not Significant" ~ "No change",
    )) 

write.csv(clade_results_df, "output_tables/Microcystis_clade_differential_abundance_results.csv", row.names = FALSE)
```


```{r}
ggplot(clade_results_df, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(shape = Month, fill = sig_label, alpha = sig_label),color = "black",stroke = 0.3, size = 4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  ggrepel::geom_text_repel(data = clade_results_df %>% 
                             filter(adj.P.Val < 0.05 & abs(logFC) > 0.1), 
                           aes(label = Clade), size = 3, max.overlaps = 8, fontface = "bold") +
  scale_alpha_manual(values = c("2018 Lake Erie" = 1, 
                                "2019 Lake Erie" = 1, 
                                "2019 Lake Huron" = 1,"No change" = 0.5))+
  scale_fill_manual(name = "Lake System", 
                    values = c("2018 Lake Erie" = "peachpuff2", 
                               "2019 Lake Erie" = "cyan4", 
                               "2019 Lake Huron" = "lightpink4",
                               "No change" = "grey80")) +
  scale_shape_manual(name = "Month", 
                     values = c("06" = 21, "07" = 22, 
                                "08" = 23,"09" = 24, "10" = 25))+
  guides(fill = guide_legend(override.aes = list(shape = 21)),
         alpha = "none",
         shape = guide_legend(override.aes = list(fill = "grey80"))) +
  labs(title = "Hierarchy of vulnerability within Microcystis",
       x = "Log2 Fold Change (Quagga vs Control)", y = "-Log10 Adjusted P-value") +
  theme_pubclean()+
  theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
          axis.line = element_line(linewidth=0, colour = "black"),
          axis.ticks = element_line(size=1, colour = "black"),
          axis.text.y = element_text(hjust = 0,size=15,color='black',face = "bold"),
          axis.text.x = element_text(hjust = 0.5,size=15,color='black',face = "bold", angle = 90),
          axis.title = element_text(size=15,color='black',face = "bold"),
          panel.background = element_rect(fill = "grey100"),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
          legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
          legend.text = element_text(size = 8),
          legend.box = "vertical",
          legend.position = "right",
          panel.grid.major = element_blank())

ggsave("figures/fig3C/microcystis_log2QC_clade_relabund_volcano_year_lake.png", width = 5.5, height = 4.5, dpi = 300)

```


# Unpalatability is a function of Microcystis composition over time

```{r}

focus_strain <- results_df %>%
  filter(significant!= "Not Significant") %>%
  group_by(seqnames, Clade) %>%
  summarise(n_sig = n()) %>%
  filter(n_sig >1 ) %>%
  arrange(Clade) %>%
  filter(!seqnames %in% c("Ma_QC_Ca_00000000_S207","Mw_MB_S_20031200_S109"))

strain_fc <- focus_strain %>%
  left_join(results_df, by = "seqnames") %>%
  left_join(strain_pcoa_scores %>%
              select(Date, V1, V2), by = "Date") %>%
  mutate(clade_strain = paste(Clade.x,seqnames,sep = " "))

ggplot(strain_fc, aes(V1, logFC)) +
  geom_point(aes(shape = Month,fill = Year_lake, alpha = sig_label),
             color = "black",stroke = 0.5, size = 4)+
  scale_alpha_manual(values = c("2018 Lake Erie" = 1, 
                                "2019 Lake Erie" = 1, 
                                "2019 Lake Huron" = 1,"No change" = 0.1))+
  scale_fill_manual(name = "Lake System", 
                    values = c("2018 Lake Erie" = "peachpuff2", 
                               "2019 Lake Erie" = "cyan4", 
                               "2019 Lake Huron" = "lightpink4",
                               "No change" = "grey80")) +
  scale_shape_manual(name = "Month", 
                     values = c("06" = 21, "07" = 22, 
                                "08" = 23,"09" = 24, "10" = 25))+
  #ggrepel::geom_text_repel(data = strain_fc %>% 
  #                           filter(adj.P.Val < 0.05) , 
  #                         aes(label = strain_clade), 
  #                         size = 2, max.overlaps = 8, fontface = "bold") +
  guides(fill = guide_legend(override.aes = list(shape = 21)),
         alpha = "none",
         shape = guide_legend(override.aes = list(fill = "grey80"))) +
  labs(title = "Vulnerability change depending on composition",
       x = "PCo1", y = "Log2 Fold Change (Quagga vs Control)") +
  facet_wrap(~clade_strain, ncol = 3,scales = "fixed") +
  theme_pubclean()+
  theme(plot.title= element_text(size=15,color='black',hjust=0.39,face = "bold"),
        axis.line = element_line(linewidth=0, colour = "black"),
        axis.ticks = element_line(size=1, colour = "black"),
        axis.text.y = element_text(size=14,color='black',face = "bold"),
        axis.text.x = element_text(vjust = 0.5,size=14,color='black',face = "bold", angle = 90),
        axis.title = element_text(size=15,color='black',face = "bold"),
        panel.background = element_rect(fill = "grey100"),
        strip.background = element_rect(fill = "grey100"),
        strip.text.x.top = element_text(size = 10, face = "bold", color = "black"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
        legend.title = element_text(size=10,color='black',hjust=0.5, face = "bold"),
        legend.text = element_text(size = 10),
        legend.box = "vertical",
        legend.position = "right",
        panel.grid.major = element_blank())

ggsave("figures/fig3C/strain_log2QC_PCo1.png", width = 9 , height = 6, dpi = 300)

```





